---
title: "RNAseq Analysis of CBFA2T3-GLIS2 Transduced Cord Blood"
author: "Jenny Smith"
date: "12/28/20"
output: html_document
---


# Set-up 

```{r setup, cache = FALSE, include = FALSE}
require(knitr)
knitr::opts_knit$set(root.dir = file.path(PROJHOME,"2017.02.15_CBF-GLIS_DEG/2020.12.23_CBFGLIS_Models"))
```

```{r}
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),
                      tidy=TRUE,
                      fig.align='center',
                      fig.width = 10, fig.height = 10)
node=Sys.info()[["nodename"]]
if(!grepl("local", node)){
  print(node)
  options(bitmapType = 'cairo')
  grDevices::X11.options(type='cairo')
}

options(stringsAsFactors = FALSE)
table = function (..., useNA = 'ifany') base::table(..., useNA = useNA)
```

```{r message = FALSE, warning=FALSE}
library(stringr)
library(magrittr)

library(ggplot2)
library(gridExtra)
library(ggpubr)
library(ggrepel)

library(dplyr)
library(tidyr)
library(tibble)


library(DeGSEA)
library(edgeR)
getwd()
```

```{r}
#hrbrthemes
# devtools::install_github("thomasp85/patchwork")
library(patchwork)
```


#Define Functions 

```{r}
kallisto_rmDups <- function(kallisto.df, geneIDmap){
        
      #Fix the PAR regions. Need to sum them since the fasta sequence is identical between PAR_Y and the X chromosome gene 
        PAR <- filter(geneIDmap, grepl("_PAR_Y", gene_id))
        PAR <- PAR %>% 
          pull(gene_id) %>% 
          gsub("_PAR_Y", "", .) %>% 
          paste(., collapse="|")
        
        #For each PAR region, use ColSums to combine them 
        idx <- grep(PAR, rownames(kallisto.df), value=TRUE) 
        length(idx) #90 PAR_Y and their X-chromosome pair
        
        pairs <- lapply(seq(2, length(idx), by=2), 
                        function(x){ 
                          rows <- idx[c((x-1):x)]
                          colSums(kallisto.df[rows,])
                        }) %>% 
          do.call(rbind, .)
        
        #Remove the PAR_Y rows and add in the summed gene expression
        non_PAR <- idx[seq(2, length(idx), by=2)-1]
        kallisto.df <- kallisto.df[!grepl("_PAR_Y$", rownames(kallisto.df)), ]
        kallisto.df[non_PAR,] <- pairs


        # dim(kallisto.df) #59808  2116
        # head(kallisto.df[,1:5])

        #convert Ensembl IDs to Gene Symbols
        kallisto.df.sym <- as.data.frame(kallisto.df) %>%
          rownames_to_column("gene_id") %>%
          left_join(., dplyr::select(geneIDmap,gene_id,gene_name),
                    by="gene_id") %>%
          dplyr::select(gene_id,gene_name, everything())

        #seperate out those genes without duplicate symbols
        kallisto.df.sym_noDups <- kallisto.df.sym %>%
          filter(!duplicated(gene_name), !duplicated(gene_name, fromLast = TRUE))

        #identify which columns are numeric counts columns
        numeric_cols <- sapply(kallisto.df.sym, function(x) is.numeric(x))
        numeric_cols <- names(numeric_cols)[which(numeric_cols)]

        #calculate the averate for all duplicate gene symbols.
        kallisto.df.sym_Dups <- kallisto.df.sym %>%
          filter(duplicated(gene_name) | duplicated(gene_name, fromLast = TRUE)) %>%
          arrange(gene_name) %>%

          #Calculate IQR to measure variance
          rowwise() %>%
          mutate(IQR=IQR(c_across(all_of(numeric_cols)))) %>%
          ungroup() %>%

          #use the max IQR expression to address duplicate gene symbols
          group_by(gene_name) %>%
          mutate(Rank=rank(IQR,ties.method= "first")) %>%
          mutate(Keep=ifelse(Rank==max(Rank), TRUE, FALSE)) %>%
          ungroup() %>%
          dplyr::select(IQR:Keep, everything())

        # head(kallisto.df.sym_Dups[,1:10])
        # dim(kallisto.df.sym_Dups) #1719 2118

        #remove duplicated gene_names
        kallisto.df.sym_rmDups <- kallisto.df.sym_Dups %>%
          filter(Keep) %>%
          dplyr::select(-IQR,-Keep, -Rank)

        #
        # dim(kallisto.df.sym_rmDups) #174 2118
        # head(kallisto.df.sym_rmDups)

        kallisto.df.sym_final <- bind_rows(kallisto.df.sym_noDups, kallisto.df.sym_rmDups)
        return(kallisto.df.sym_final)
        # dim(kallisto.df.sym_final) #58263  2118
        # head(kallisto.df.sym_final[,1:5])
}
```

```{r}
#Custom script to plot the experimental RNAseq expression with the patient samples. 
scatter_comparison <- function(df,cols, genes){

  expn <- logCPM.cbf[genes, df$Sample] %>% 
      as.data.frame() %>% 
      rownames_to_column("gene") %>% 
      gather(Sample, log2CPM, -gene) %>% 
      left_join(., dplyr::select(df, Comparisons, Sample), 
                by="Sample")  %>% 
      
      group_by(Comparisons, gene) %>% 
      mutate(Mean_Expression=mean(log2CPM)) %>% 
      ungroup() %>% 
      
      dplyr::select(gene, Comparisons, Mean_Expression) %>% 
      distinct() %>% 
      
      spread(Comparisons, Mean_Expression)
    

    col_x <- cols[1]
    col_y <- cols[2]
    form <- as.formula(paste(col_y, " ~ 0+", col_x))
    reg <- lm(form, data = expn)
    s <- summary(reg)
    
    
    # plot(reg)
    p.val <- round(s$coefficients[4],digits=3)
    p.val <- ifelse(p.val == 0, "p<0.001",paste0("p=",p.val))
    
    y=quantile(expn[[col_x]])[5]
    x=quantile(expn[[col_y]])[2]
    
    scatter <- ggplot(expn, aes_string(x=col_x, y=col_y)) +
      geom_point(color="red4", alpha=0.5) +
      geom_smooth(method="lm",formula='y ~ x',  color="black", linetype=2) +
      annotate(geom="text", x = x, y=y, label=p.val, size=6) +
      labs(title="Mean Expression of Common Dysregulated Genes") +
      theme_classic() +
      theme(text=element_text(size=14))
  
  
  return(scatter)
}

```


```{r fig.width=12}
make_umap_2d <- function(umap_res.df, colors_list,dims){
  
  
       subset <- filter(umap_res.df,grepl("CordBlood", AML_Subtype)) %>% 
       mutate_at(vars(AML_Subtype), ~factor(.,levels=c("EC_Week1_GFP_CordBlood",
                                                          "EC_Week1_CBFGLIS_CordBlood",
                                                          "EC_Week6_CBFGLIS_CordBlood",
                                                          "EC_Week12_CBFGLIS_CordBlood",
                                                          "Myeloid_Week1_GFP_CordBlood",
                                                          "Myeloid_Week1_CBFGLIS_CordBlood",
                                                          "Myeloid_Week6_CBFGLIS_CordBlood")))
        models <- unique(pull(subset, AML_Subtype))  
        models <- models[order(models)]
        shapes <-   case_when(grepl("Week1_", models) ~ 21,
                    grepl("Week6",models) ~  24,
                    grepl("Week12", models) ~ 22) %>% 
          set_names(models)
        
        bulk_aml <- filter(umap_res.df,!grepl("CordBlood", AML_Subtype))
        var1 <- rlang::sym(dims[1])
        var2 <- rlang::sym(dims[2])
        
        group_labs <- bulk_aml %>% 
          select(x,y,z, AML_Subtype) %>% 
          group_by(AML_Subtype) %>% 
          mutate(pos1 := quantile(!!var1, probs=0.85),
                 pos2 := quantile(!!var2,probs=0.25)) %>% 
          # mutate(pos1 := mean(!!var1),
          #        pos2 := mean(!!var2)) %>% 
          ungroup() %>% 
          select(pos1,pos2,AML_Subtype) %>% 
          distinct() %>% 
          left_join(., rownames_to_column(data.frame(colors=cc_newData$AML_Subtype)),
                    by=c("AML_Subtype"="rowname"))
        
        labs <- case_when(
          all(dims %in% c("x","y")) ~ c("UMAP_1","UMAP_2"),
          all(dims %in% c("x","z")) ~ c("UMAP_1","UMAP_3"),
          all(dims %in% c("y","z")) ~ c("UMAP_2","UMAP_3"))    
        
        pos <- ifelse(all(dims %in% c("x","y")), "left", "none")
        
        ggplot() +
          geom_point(data=bulk_aml,
                     mapping=aes_string(x=dims[1],y=dims[2], color="AML_Subtype", group="AML_Subtype"),
                     shape=19,
                     size=bulk_aml$pt_size,
                     alpha=0.6) +
          geom_point(data=subset,
                     mapping=aes_string(x=dims[1],y=dims[2],
                                 fill="AML_Subtype",
                                 shape="Time_Point_Group"),
                     size=3,
                     alpha=0.7) +
          scale_color_manual(values=colors_list$AML_Subtype) +
          scale_fill_manual(values=colors_list$AML_Subtype) +
          scale_shape_manual(values=c("Week1"=21, "Week6"=24, "Week12"=22)) +
          guides(fill=guide_legend(override.aes = list(size =5, 
                                                       color=colors_list$AML_Subtype[grep("Cord", names(colors_list$AML_Subtype))],
                                                       shape=shapes,
                                                       alpha=1.0)),
                 shape=guide_legend(override.aes = list(size=5)),
                 color=guide_legend(override.aes = list(size=5,
                                                        alpha=1.0))) +
          labs(x=labs[1], y=labs[2]) +
          theme_classic() +
          geom_text_repel(data = group_labs, 
                          mapping = aes(x=pos1, y=pos2, label=AML_Subtype),
                          color=group_labs$colors,
                          # nudge_x=c(2.0),
                          # position = position_dodge(width = 2),  
                          # segment.size=10.0,
                          # min.segment.length=unit(5, "mm")
                          ) +
          theme(axis.text = element_text(size=16, color="black"), 
                legend.position = pos,
                axis.title = element_text(size=18, color="black")) 

        
}
```

```{r}
make_facet_boxplots <- function(genes.of.interest, breaks){
  
  upper_lim <- ifelse(max(breaks) > max(genes.of.interest$CPM),  max(breaks), max(genes.of.interest$CPM))
  
  # breaks <- plyr::round_any(seq(0,m, length.out = 5), accuracy=10)
  ggplot(data=genes.of.interest, aes(x=Comparisons, y=CPM, fill=Comparisons)) +
      geom_boxplot(data=filter(genes.of.interest,Facet=="CBFA2T3-GLIS2 HSPCs"),
                   mapping=aes(x=Comparisons, y=CPM, fill=Comparisons, color=Comparisons),
                  alpha=0.2,outlier.shape = NA,  inherit.aes = TRUE) +
      geom_beeswarm(data=filter(genes.of.interest,Facet=="CBFA2T3-GLIS2 HSPCs"),
                    mapping=aes(x=Comparisons, y=CPM, fill=Comparisons,  color=Comparisons, shape=Time_Point_Group),
                    size=3, alpha=0.75,
                    inherit.aes = TRUE) +


      geom_boxplot(data=filter(genes.of.interest,Facet != "CBFA2T3-GLIS2 HSPCs"),
                   mapping=aes(x=Comparisons, y=CPM,  fill=Comparisons,color=Comparisons),
                   alpha=0.25,
                   outlier.shape = NA,
                   inherit.aes = TRUE) +
      geom_beeswarm(data=filter(genes.of.interest,Facet != "CBFA2T3-GLIS2 HSPCs",
                                !c(gene == "BMP2" & Group=="NBM")),
                   mapping=aes(x=Comparisons, y=CPM, fill=Comparisons,color=Comparisons,shape=Time_Point_Group),
                  alpha=0.4,
                  size=1.5,
                  inherit.aes = TRUE) +
      facet_wrap(~gene+Facet, ncol=6,scales = "free_x") +


      scale_fill_manual(values=cc.cultures) +
      scale_color_manual(values=cc.cultures) +
      scale_shape_manual(values=c("Week1"=21, "Week6"=24, "Week12"=22,
                                  "NBM"=19, "CBFA2T3-GLIS2"=19)) +

      guides(fill=guide_legend(ncol=3),color=guide_legend(ncol=3), shape="none") +
      scale_y_continuous(breaks = breaks, limits=c(0,upper_lim)) +

      labs(x="",y="Expression (CPM)") +
      theme_classic() +
      theme(axis.title = element_text(size=14),
            axis.text.y = element_text(color="black", size=14),
            axis.text.x = element_text(angle=25, vjust=1, hjust=1, size=11,color="black"),
            panel.border = element_rect(color="black", fill=NA),
            plot.margin = margin(l=4, unit="cm"),
            strip.text = element_text(size=14),
            legend.text = element_text(size=14),
            legend.position = "none")
  
  
}
```


```{r fig.width=10, fig.height=6}
make_gsva_facet_boxplots <- function(gsva_res, breaks){
  
  
  # breaks <- plyr::round_any(seq(min(gsva_res$Score), max(gsva_res$Score), length.out = 5), accuracy=0.05)
  start <- ifelse(min(gsva_res$Score) < min(breaks),min(gsva_res$Score), min(breaks))
  end <- ifelse(max(gsva_res$Score) > max(breaks),max(gsva_res$Score), max(breaks))
  
  ggplot(data=gsva_res, aes(x=Comparisons, y=Score, fill=Comparisons)) +
      geom_boxplot(data=filter(gsva_res,Facet=="CBFA2T3-GLIS2 HSPCs"),
                   mapping=aes(x=Comparisons, y=Score, fill=Comparisons, color=Comparisons),
                  alpha=0.2,outlier.shape = NA,  inherit.aes = TRUE) +
      ggbeeswarm::geom_beeswarm(data=filter(gsva_res,Facet=="CBFA2T3-GLIS2 HSPCs"),
                    mapping=aes(x=Comparisons, y=Score, fill=Comparisons,  color=Comparisons, shape=Time_Point_Group),
                    size=3,  
                    alpha=0.75,
                    inherit.aes = TRUE) +


      geom_boxplot(data=filter(gsva_res,Facet != "CBFA2T3-GLIS2 HSPCs"),
                   mapping=aes(x=Comparisons, y=Score,  fill=Comparisons,color=Comparisons),
                   alpha=0.25,
                   outlier.shape = NA,
                   inherit.aes = TRUE) +
      ggbeeswarm::geom_beeswarm(data=filter(gsva_res,Facet != "CBFA2T3-GLIS2 HSPCs"),
                   mapping=aes(x=Comparisons, y=Score, fill=Comparisons,color=Comparisons,shape=Time_Point_Group),
                  alpha=0.4,
                  size=1.5,
                  inherit.aes = TRUE) +
      facet_wrap(~path+Facet, ncol=2,scales = "free_x") +
      


      scale_fill_manual(values=cc.cultures) +
      scale_color_manual(values=cc.cultures) +
      scale_shape_manual(values=c("Week1"=21, "Week6"=24, "Week12"=22,
                                  "NBM"=19, "CBFA2T3-GLIS2"=19)) +

      guides(fill=guide_legend(ncol=3),color=guide_legend(ncol=3), shape="none") +
      scale_y_continuous(breaks = breaks, limits=c(start,end)) + #limits=c(min(breaks), max(breaks))

      labs(x="",y="Enrichment Score") +
      theme_classic() +
      theme(axis.title = element_text(size=14),
            axis.text.y = element_text(color="black", size=14),
            axis.text.x = element_text(angle=25, vjust=1, hjust=1, size=11,color="black"),
            panel.border = element_rect(color="black", fill=NA),
            plot.margin = margin(l=4, unit="cm"),
            strip.text = element_text(size=14),
            legend.text = element_text(size=14),
            legend.position = "bottom")
  
  
}
```

```{r}
# https://github.com/gaospecial/ggVennDiagram/blob/master/R/multi_implement.R
multi_union <- function ( ...,l = NULL){
  if (is.null(l)) l <- list(...)
  n <- length(l)
  if (n < 3){
    stop("At least three parameters.")
  }
  else {
    v <- l[[1]]
    for (i in 2:n){
      v <- union(v,l[[i]])
    }
    return(v)
  }
}

#' @rdname multi
multi_intersect <- function ( ...,l = NULL){
  if (is.null(l)) l <- list(...)
  n <- length(l)
  if (n < 3){
    stop("At least three parameters.")
  }
  else {
    v <- l[[1]]
    for (i in 2:n){
      v <- intersect(v,l[[i]])
    }
    return(v)
  }
}


#' @rdname multi
multi_setdiff <- function(...,l=NULL){
  if (is.null(l)) l <- list(...)
  n <- length(l)
  if (n < 3){
    stop("At least three parameters.")
  }
  else {
    v <- l[[1]]
    for (i in 2:n){
      v <- setdiff(v,l[[i]])
    }
    return(v)
  }
}
```


# ClinData

```{r}
merged <- read.csv(file.path(CDE,"Merged/TARGET_AML_0531_1031_merged_CDEs_05.21.21.csv"),
                   na.strings = c("#N/A", "NA", ".",""))

merged <- merged %>% 
  dplyr::filter(!grepl("Unknown", USI)) %>%
  dplyr::filter(Eligibility_Comments != "remove")


dim(merged) #2217  150
```

```{r}
sample_info <- read.csv(file.path(TARGET,
                                    "SequencingDataMatrix/TARGET_AML_Ribodepleted_Manifest_08.12.21.csv")) 

dim(sample_info)
table(sample_info$Batch)
```


```{r}
#Create an excel workbook to save the raw data used to generate the figures. 
supp_table <- openxlsx::createWorkbook()


# openxlsx::saveWorkbook(supp_table,
#                        "CBFA2T3.GLIS2_HSPCs_Supplemental_Tables_Raw_Figure_Data.xlsx",
#                        overwrite = T)
```


# Gene Annotations

```{r}
geneIDmap <- read.delim(file.path(PROJHOME,"0000.00.02_Reference_GeneInfo/gencode.v29_RepBase.v24.01_GeneLevel_IDmap_anno_5.14.21.txt"))

dim(geneIDmap) #59853    23
# View(geneIDmap)
```

```{r}
CBFGLISvsAML <- read.csv(file.path(PROJHOME,"0000.00.05_DEG_Lists/GRCh38_hg38/Kallisto/CBFA2T3GLIS2vsOthers_DEGs.csv"))

dim(CBFGLISvsAML)
# head(CBFGLISvsAML, n=10)
```

```{r}
CBFGLISvsNBM <- read.csv(file.path(PROJHOME, "0000.00.05_DEG_Lists/GRCh38_hg38/Kallisto/CBFA2T3GLIS2vsNBM_DEGs.csv"))

dim(CBFGLISvsNBM)
# head(CBFGLISvsNBM, n=10)
```

```{r}
intersection <- CBFGLISvsAML %>% 
  select(gene_id=gene, gene_name, logFC, adj.P.Val) %>% 
  rename_at(vars(logFC:adj.P.Val), ~paste0(.,"_CBFGLISvsAML")) %>% 
  inner_join(., select(CBFGLISvsNBM, 
                       gene_id=gene, logFC, adj.P.Val) %>% 
               rename_at(vars(logFC:adj.P.Val), ~paste0(.,"_CBFGLISvsNBM")),
             by="gene_id")


# intersection
# dim(intersection)
# write.csv(intersection, "TARGET_AML_CBFGLISpatients_vsAML_vsNBM_intersection_DEGs.csv", row.names = FALSE)
```


# Raw Counts

```{r}
cts <- readRDS("Expression_Data/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Models_Kallisto_Quant_GeneLevel_scaledTPM_counts.RDS")
colnames(cts)[grep("PATGIG|PATISD", colnames(cts))] <- gsub("_replicate", "", grep("PATGIG|PATISD", colnames(cts), value=T))

geneIDs <- cts[,1:2]
rownames(cts) <- cts$gene_name
cts <- cts[,-c(1:2)]

dim(cts)
# head(cts[,1:5]) #58261  2166
```

```{r}
TPM <- readRDS("Expression_Data/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Models_Kallisto_Quant_GeneLevel_Abundance_TPM.RDS")
colnames(TPM)[grep("PATGIG|PATISD", colnames(TPM))] <- gsub("_replicate", "", grep("PATGIG|PATISD", colnames(TPM), value=T))


geneIDs.tpm <- TPM[,1:2]
rownames(TPM) <- TPM$gene_name
TPM <- TPM[,-c(1:2)]


dim(TPM)
# head(TPM[,1:5])
```


## Remove Duplicate Genes

```{r eval=FALSE}
cts.otherAML <- readRDS(file.path(PROJHOME, "0000.00.03_ExpressionMatrices/Kallisto_GRCh38_Gencode_v29/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Kallisto_Quant_GeneLevel_scaledTPM_counts.RDS"))


# head(cts.otherAML[,1:5])
# dim(cts.otherAML) #59853  2116

cts.CBFGLIS <- readRDS("Expression_Data/TARGET_AML_CBFGLIS_Models_geneLevel_scaledTPM_counts.RDS")
rownames(cts.CBFGLIS) <- gsub("\\.[0-9]{1,2}", "", rownames(cts.CBFGLIS))

# 
# head(cts.CBFGLIS[,1:5])
# dim(cts.CBFGLIS) #58263    50


identical(rownames(cts.CBFGLIS), rownames(cts.otherAML)) #TRUE
table(rownames(cts.CBFGLIS) %in% rownames(cts.otherAML)) #TRUE
# grep("^[0-9]{1,2}\\-", rownames(cts.CBFGLIS), value=T) #OK no  gene names edited by excel. 
# table(rownames(cts) %in% rownames(cts.CBFGLIS)) #TRUE
# any(colnames(cts.CBFGLIS) %in% colnames(cts.otherAML)) #OK FALSE

cts.geneLevel <- cbind(cts.otherAML, cts.CBFGLIS)
# dim(cts.geneLevel) #59853  2166
# head(cts.geneLevel[,1:5])

cts <- kallisto_rmDups(cts.geneLevel, geneIDmap = geneIDmap)

head(cts[,1:5])
dim(cts) #58263  2168 (it changed bc the damn last version I used of the geneIDmap had dates ughhhh)


# saveRDS(cts,"Expression_Data/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Models_Kallisto_Quant_GeneLevel_scaledTPM_counts.RDS")
```

```{r eval=FALSE}
TPM.otherAML <- readRDS(file.path(PROJHOME, "0000.00.03_ExpressionMatrices/Kallisto_GRCh38_Gencode_v29/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Kallisto_Quant_GeneLevel_Abundance_TPM.RDS"))


# head(TPM.otherAML[,1:5])
# dim(TPM.otherAML) #59853  2116

TPM.CBFGLIS <- readRDS("Expression_Data/TARGET_AML_CBFGLIS_Models_geneLevel_TPM_abundance.RDS")
rownames(TPM.CBFGLIS) <- gsub("\\.[0-9]{1,2}", "", rownames(TPM.CBFGLIS))

# 
# head(TPM.CBFGLIS[,1:5])
# dim(TPM.CBFGLIS) #59853    50


identical(rownames(TPM.CBFGLIS), rownames(TPM.otherAML)) #TRUE
table(rownames(TPM.CBFGLIS) %in% rownames(TPM.otherAML)) #TRUE
# grep("^[0-9]{1,2}\\-", rownames(TPM.CBFGLIS), value=T) #OK no  gene names edited by excel.
# table(rownames(TPM) %in% rownames(TPM.CBFGLIS)) #TRUE
# any(colnames(TPM.CBFGLIS) %in% colnames(TPM.otherAML)) #OK FALSE

# TPM.geneLevel <- cbind(TPM.otherAML, TPM.CBFGLIS)
# dim(TPM.geneLevel) #59853  2166
# head(TPM.geneLevel[,1:5])

TPM <- kallisto_rmDups(TPM.geneLevel, geneIDmap = geneIDmap)

# head(TPM[,1:5])
# dim(TPM) #58263  2168 


# saveRDS(TPM,"Expression_Data/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Models_Kallisto_Quant_GeneLevel_Abundance_TPM.RDS")
```

# Define Samples and Contrasts 

```{r}
CBFGLIS.models <- read.csv("TARGET_AML_CBFGLIS_CordBlood_Models_Sample_Conditions.csv")

# head(CBFGLIS.models)
dim(CBFGLIS.models) #50 56
# View(CBFGLIS.models)Ω
```

```{r}
CBFGLIS_samples <- CBFGLIS.models %>%  
  filter(grepl("CordBlood", AML_Subtype),
         !grepl("sort", Sample),
         !grepl("TARGET.20.CB34pos.Dayminus1.00.01R", Sample)) %>%
  
  mutate_at(vars(Time_Point_Group), ~case_when(
    .=="Week7" ~ gsub("7", "6", .),
    TRUE ~ .)) %>% 
  bind_rows(., filter(sample_info, grepl("CD34|NBM|GLIS2",AML_Subtype),
                      grepl("diagnostic|NBM|CD34_PB", Time_point))) %>%
  left_join(.,dplyr::select(merged, USI, M7_AML,
                     BM.blasts=Bone.marrow.leukemic.blast.percentage....),
            by="USI") %>%
  mutate_at(vars(c("Time_point","Time_Point_Group","Culture_condition",
                   "Transduction_condition","M7_AML")),
            ~case_when(
              is.na(.) ~ AML_Subtype,
              TRUE ~ .)) %>%
  
  mutate(Comparisons=paste(Culture_condition, Time_Point_Group, AML_Subtype, sep="_")) %>%
  mutate(Comparisons=gsub("^(CBFA2T3.GLIS2|NBM|CD34_PB)_.+", "\\1", Comparisons)) %>% 
  
  mutate(USI1=USI,
         USI=Sample) %>% 
  arrange(AML_Subtype) %>% 
  set_rownames(.$Sample)



dim(CBFGLIS_samples) #182  64
table(CBFGLIS_samples$Comparisons)
# colnames(CBFGLIS_samples)[colnames(CBFGLIS_samples) %in% colnames(sample_info)]
```


```{r}
samps_all <- sample_info %>% 
  filter(grepl("^AML|CD34_PB|NBM|Normal_PB|^FlowSorted", Group))  %>% 
  mutate_at(vars(Time_point), ~case_when(
    Group=="FlowSorted" & grepl("Unsorted.0[39]A",Sample) ~ "diagnostic",
    TRUE ~ .)) %>% 
  mutate_at(vars(Sample), ~case_when(
    grepl("PATGIG|PATISD", .) ~ gsub("_replicate", "", .),
    TRUE ~ .)) %>% 
  filter(grepl("diagnostic|CD34_PB|NBM|Normal_PB", Time_point)) %>% 
  
  #Keep only the CBFGLIS samples of interest from EC and Myeloid cultures
  bind_rows(., filter(CBFGLIS_samples, grepl("^EC|^Myeloid", Comparisons), 
                      !grepl("Myeloid_Week12|EC_Week9|Myeloid_Week6_GFP_CordBlood", Comparisons),
                      !grepl("mock", Sample))) %>% 
  mutate_at(vars(Time_Point_Group), ~case_when(
    is.na(.) ~ Time_point,
    TRUE ~ .)) %>% 
  filter(Sample %in% colnames(cts),
         !grepl("_replicate", Sample)) %>% 
  set_rownames(.$Sample)
  


dim(samps_all)
table(samps_all$Group)
table(samps_all$Time_Point_Group)
table(samps_all$Comparisons)


# table(CBFGLIS_samples$Sample  %in% samps_all$Sample)
```





# Subset Counts Data

```{r}
in_counts <- cts[,CBFGLIS_samples$Sample]

dim(in_counts) #58263   182
head(in_counts[,1:5])

AML <- ! grepl("BM[0-9]|R[O0][0-9]", colnames(in_counts))
keep <- rowSums(cpm(in_counts[,AML]) >= 1) >= 0.05*ncol(in_counts[,AML])
cts.filtered <- in_counts[keep, ]

dge <- DGEList(counts=cts.filtered)
dge <- calcNormFactors(dge,method = "TMMwsp")

logCPM <- edgeR::cpm(dge,log=TRUE,normalized.lib.sizes=TRUE, prior.count=1)
CPM <- edgeR::cpm(dge,log=FALSE,normalized.lib.sizes=TRUE, prior.count=1)

dim(logCPM)  #25665   182
head(logCPM[,1:5])
```

```{r}
genes <- grep("^ENSG", geneIDs$gene_id)

cts_all <- cts[genes, samps_all$Sample]
AML <- !grepl("BM[0-9]|R[O0][0-9]", colnames(cts_all))
keep <- rowSums(cpm(cts_all[,AML]) >= 1) >= 0.025*ncol(cts_all[,AML])

#Filter to remove low/zero count genes
cts_all <- cts_all[keep, ]
dge_all <- DGEList(counts=cts_all)
dge_all <- calcNormFactors(dge_all,method = "TMMwsp")

logCPM_all <- edgeR::cpm(dge_all,log=TRUE,normalized.lib.sizes=TRUE, prior.count=1)
# dim(cts_all) #27,245  1618

head(logCPM_all[,1:5])
```

# Unsupervised Clustering

*unsupervised clustering first
PCA/MDS/UMAP

samples correlations with heirarchal clustering

*heatmap with DEGs from CBFGLIS samples vs OtherAML 
- include D7, ~D45, D84 and GFP samples 
- global view of which blocks of genes are shared. 

```{r}
library(DelayedArray)

# Mean vs Dispersion Feature Selection 
obj <- seqGlue::calc_dispersion(as.matrix(CPM), removeOutliers = TRUE) #removes outlier genes/transcripts based on cooks distance
gc()

sg_all <- seqGlue::get_selected_genes(seqGlue::select_genes(obj, top_n=NULL))
length(sg_all) #4890 genes 
# head(sg_all)
```


## PCA

```{r}
# cc <- list("fill"=,"color"=)
```

```{r}
p <- pull(CBFGLIS.models, AML_Subtype) %>% 
  set_names(CBFGLIS.models$Sample)

PCA_models <- PCA(expnData = cts.filtered,
                  phenovector = p,
                  title = "PCA Analysis with Transduced Cord Blood Models", 
                  ntop = 1000,
                  round = T,PC3=TRUE)
```

```{r fig.height=5, fig.width=15 }
pm.1 <- PCA_models$pca_plot + scale_color_brewer(palette = "Paired")
pm.2 <- PCA_models$pca_plot2 + scale_color_brewer(palette = "Paired")

# pdf("Figures/CBFGLIS_Models_Controls_Top1000_mostVariedGenes_PCA_plot.pdf", height = 5, width = 15)
pm.1 + pm.2
# dev.off()
```


```{r}
p <- pull(CBFGLIS_samples, AML_Subtype) %>% 
  set_names(CBFGLIS_samples$Sample)

PCA_simple <- PCA(expnData = cts.filtered,
                  phenovector = p,
                  title = "PCA Analysis with Patients and Cord Blood Models", 
                  ntop = 1000,
                  round = T,PC3=TRUE)
```

```{r fig.height=5, fig.width=15}
ps.1 <- PCA_simple$pca_plot + scale_color_brewer(palette = "Paired")
ps.2 <- PCA_simple$pca_plot2 + scale_color_brewer(palette = "Paired")

# pdf("Figures/CBFGLIS_Pts_Models_Controls_Top1000_mostVariedGenes_PCA_plot.pdf", height = 5, width = 15)
ps.1 + ps.2
# dev.off()
```

```{r}
class(PCA_simple$vst)
dim(assay(PCA_simple$vst))

table(rownames(assay(PCA_simple$vst)) %in% sg_all)
```

```{r}
PCA_res <- pca_custom(expnData = assay(PCA_simple$vst)[sg_all, CBFGLIS_samples$Sample],
                      CDE=CBFGLIS_samples,
                      fillCol = "AML_Subtype",
                      colorCol = "Time_point",
                      PC3 = TRUE)
```

```{r fig.height=7, fig.width=18}
pca.p1 <- PCA_res$plot.1 +
  scale_fill_brewer(palette = "Paired") +
  scale_color_brewer(palette = "Set3")

pca.p2 <- PCA_res$plot.2 +
  scale_fill_brewer(palette = "Paired") +
  scale_color_brewer(palette = "Set3")

# pdf("Figures/CBFGLIS_Pts_Models_Controls_Condition_TimePoint_PCA_plot.pdf", height=7, width = 18)
pca.p1 + pca.p2
# dev.off()
```

## Heatmap

Heat map of CBFGLIS genes showing clustering of the primary NBM samples, primary CBF AML; GFP EC wk 1; CBF EC wk 1, 7(day 42+day45), 12 (day 84+day87); GFP MC wk1; and CBF MC wk 1, 6, 12.

```{r}
heatmap.samps <- CBFGLIS_samples %>% 
  filter(!grepl("pool", Sample)) %>% 
  #CD34_PB
  filter(grepl("CBFA2T3|NBM|EC_Week1_GFP|EC_Week(1|6|12)_CBFGLIS|Myeloid_Week1_GFP|Myeloid_Week(1|6)_CBFGLIS",
               Comparisons)) %>% 
  filter(!c(Group=="NBM" & grepl("_rep", Sample))) %>% 
  mutate(Comparisons=factor(Comparisons, levels=c(
                                                  # "CD34_PB", 
                                                  "NBM","CBFA2T3-GLIS2",
                                                    "EC_Week1_GFP_CordBlood",
                                                    "EC_Week1_CBFGLIS_CordBlood",
                                                    "EC_Week6_CBFGLIS_CordBlood",
                                                    "EC_Week12_CBFGLIS_CordBlood",
                                                    "Myeloid_Week1_GFP_CordBlood",
                                                    "Myeloid_Week1_CBFGLIS_CordBlood",
                                                    "Myeloid_Week6_CBFGLIS_CordBlood"))) %>%
  mutate(Labels=gsub("_CordBlood", "", Comparisons),
         Time_Point_Group=factor(ifelse(grepl("NBM|CBF", Time_Point_Group), "None", Time_Point_Group),
                                 levels = c("Week1","Week6", "Week12","None"))) %>% 
  arrange(desc(Comparisons)) %>%
  set_rownames(.$Sample)
  
# levels(heatmap.samps$Comparisons)
table(heatmap.samps$Labels)


Cols <- c("Time_Point_Group","Labels")
# cc_heatmaps <- colorCodes_aheatmap(df=select(heatmap.samps,all_of(Cols)))
cc_heatmaps <- list()
cc_heatmaps[["Time_Point_Group"]] <- c("Week1" = "skyblue",
                 "Week6" = "dodgerblue2",
                 "Week12" = "navy",
                 "None" = "white")
# cc_heatmaps[["Culture_condition"]] <- c("CBFA2T3-GLIS2"="mediumorchid2", 
#                                         "EC"="blue2", 
#                                         "Myeloid"="deeppink",
#                                         "NBM"="grey20")
cc_heatmaps[["Labels"]] <- c("CBFA2T3-GLIS2"="mediumorchid2", 
                                       "NBM"="grey20",
                                       "EC_Week1_GFP"="springgreen4",
                                       "EC_Week1_CBFGLIS"="blue2",
                                       "EC_Week6_CBFGLIS"="blue2",
                                       "EC_Week12_CBFGLIS"="blue2", 
                                       "Myeloid_Week1_GFP"="orange",
                                       "Myeloid_Week1_CBFGLIS"="deeppink",
                                       "Myeloid_Week6_CBFGLIS"="deeppink")
# cc_heatmaps
# View(heatmap.samps)
table(heatmap.samps$Time_Point_Group)
```

```{r}
high.FC <- CBFGLISvsAML %>%  
  filter(abs(logFC) >= quantile(abs(logFC))[4]) %>%   #75th percentile
  filter(adj.P.Val < 0.001)

p <- pull(heatmap.samps, Comparisons) %>% set_names(heatmap.samps$Sample)
g <- intersect(rownames(in_counts), high.FC$gene_name)
length(g) #1116

anno <- DeGSEA::create_HA_Labs_Hmap(expn=in_counts[,heatmap.samps$Sample],
                                    geneList = g,
                                    cc = cc_heatmaps, 
                                    CDE = heatmap.samps,
                                    cols = names(cc_heatmaps))
anno
```


```{r}
make_TMMCPM <- function(expnData, genelist,percent=0.01,add.count=0.01){
    AML <- ! grepl("BM[0-9]|R[0O][0-9]", colnames(expnData))
    AMLsamples <- ncol(expnData[,AML])

    dge <- DGEList(counts = expnData)
    #updated on 7/26/18 - really only changes the heatmap coloring since the genes for each heatmap are already known to pass this threshold.
    keep.dge <- rowSums(cpm(dge)[,AML] >= 1) >= max(2,(percent*AMLsamples)) #X% of AML samples has cpm of at least 1 for a gene. This should help if looking for genes that are low or absent in NBMs vs AMLs
    keep.dge[names(keep.dge) %in% genelist] <- TRUE
    
    dge <- dge[keep.dge,] #subset for those genes with cmp >= 1 per gene in AML samples
    dge <- calcNormFactors(dge)

    TMMCPM <- cpm(dge, normalized.lib.sizes = TRUE) #log = TRUE, prior.count = add.count
    
    #Use +1 to avoid outliers.... not necessarily true will revisit this. Using 0.01 actually cleans up the rare fusions clustering. 
    TMMCPM <- as.data.frame(apply(TMMCPM, 2, function(x) log2(x + add.count)))
}
```


```{r}
#oof this had a filter step that I did not remember. So ofc, I am missing 36 genes from the TMMCPM matrix output. so now I need to fix that. 

input <-make_TMMCPM(expnData = in_counts[,heatmap.samps$Sample], genelist = g)

dim(input) 

d <- dge_dendrograms(expnData = input[g,heatmap.samps$Sample],
                pheno = p,
                method = "ward.D2",
                log = TRUE,
                createDGE = FALSE,
                filterTopGenes = FALSE,
                genelist = NULL)

# length(g)
# table(rownames(in_counts) %in% g)
# table(rownames(input) %in% g)
# dim(d$TMMCPM)
```

```{r}
sig.genes <- high.FC %>% 
  filter(gene_name %in% g) %>% 
  group_by(gene_name) %>% 
  filter(adj.P.Val == min(adj.P.Val)) %>% 
  filter(!duplicated(gene_name)) %>% 
  ungroup()


dim(sig.genes)

# table(sig.genes$logFC > 0)
# write.csv(sig.genes, "TARGET_AML_CBFGLIS_vs_AML_75thPercentile_AbsoluteLogFC_Signature_Geneset.csv", row.names = FALSE)

gene_set_genes <- openxlsx::read.xlsx("Manuscript/Raw_Data/TARGET_AML_CBFGLIS_MS_GSEA_Cust_Gene_Sets_AL.xlsx",
                                      sheet="CG Signature (gene set)")


#write the CBFGLISvsAML DEGs with the signature genes noted:
CBFGLISvsAML <- CBFGLISvsAML %>% 
  mutate(CBFGLIS_Signature_GeneSet_ExtendedDataFigure6b=case_when(
    gene %in% sig.genes$gene ~ "Signature Gene",
    TRUE ~ "")) %>% 
  mutate(CBFGLIS_Signature_GeneSet_Figure2j=case_when(
    gene %in% gene_set_genes$gene ~ "Included in Geneset",
    TRUE ~ ""))

table(CBFGLISvsAML$CBFGLIS_Signature_GeneSet_ExtendedDataFigure6b)
table(CBFGLISvsAML$CBFGLIS_Signature_GeneSet_Figure2j)


# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFA2T3-GLIS2 DEGs")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFA2T3-GLIS2 DEGs",
#                     CBFGLISvsAML,
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```



```{r fig.width=20, fig.height=8}
#The order of the dendrogram clusters changed (why I am unsure) so I need to specify the order. 
ord <- c(ncol(d$TMMCPM):1)
t <- dendextend::rotate(as.dendrogram(d$samp.c1),  order=ord)

#The heatmap dendrogram rotation changes since I last ran it - but results are the same. 
mat_temp <- matrix(ncol=ncol(d$TMMCPM), nrow=1, dimnames = list("", labels(t))) %>%
  as.data.frame() %>%
  select(TARGET.00.RO02496.14A.01R:TARGET.20.D7.Myeloid.CBFA2T3_GLIS2.2.00.01R,
         TARGET.20.PAYGWX.Unsorted.09A.01R:TARGET.20.PATHVG.03A.01R,
         TARGET.20.PAVPKA.09A.01R:TARGET.20.D49.Myeloid.CBFA2T3_GLIS2.2.00.01R,
         TARGET.20.D45.EC.CBFA2T3_GLIS2.3.00.01R:TARGET.20.D45.EC.CBFA2T3_GLIS2.1.00.01R,
         TARGET.20.D87.EC.CBFA2T3_GLIS2.scbulkleftovers.00.01R:TARGET.20.D84.EC.CBFA2T3_GLIS2.2.00.01R)
#   # select(TARGET.00.RO02580.14A.01R:TARGET.20.D7.Myeloid.CBFA2T3_GLIS2.2.00.01R,
#   #        TARGET.20.PAYGWX.Unsorted.09A.01R:TARGET.20.PAWYZR.09A.01R,
#   #        TARGET.20.PAVPKA.09A.01R:TARGET.20.D49.Myeloid.CBFA2T3_GLIS2.2.00.01R,
#   #        TARGET.20.D45.EC.CBFA2T3_GLIS2.3.00.01R:TARGET.20.D45.EC.CBFA2T3_GLIS2.1.00.01R,
#   #        TARGET.20.D87.EC.CBFA2T3_GLIS2.scbulkleftovers.00.01R:TARGET.20.D84.EC.CBFA2T3_GLIS2.2.00.01R)


par(mar=c(25,4,10,0), cex=0.8)
plot(dendextend::rotate(as.dendrogram(d$samp.c1),  order=ord))
plot(dendextend::rotate(as.dendrogram(d$samp.c1),  order=colnames(mat_temp)))

# View(colnames(mat_temp))

```


```{r fig.height=10, fig.width=20}
# pdf("Figures/CBFGLIS_Pts_Models_Controls_DEGsvsOtherAML_75thPercentileFC_Heatmap_7.23.21.pdf", height = 10, width=20)

HA_AML <- ComplexHmap(mat = d$TMMCPM, #logCPM[DEGs.AML, heatmap.samps$Sample]
                      name = "Z-Scores",
                      scale=TRUE,
                      dge_dendrograms.res = d,
                      samp_dend_order = colnames(mat_temp), #colnames(mat_temp)
                      hmap_anno_obj = anno$annoColumn)

HA_AML
# dev.off()
```

```{r}
# dim(HA_AML@matrix)
# dim(d$TMMCPM)

# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFA2T3-GLIS2 Heatmap Z-Scores")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFA2T3-GLIS2 Heatmap Z-Scores",
#                     HA_AML@matrix,
#                     keepNA=FALSE,
#                     rowNames=TRUE)

```

```{r fig.height=10, fig.width=20}
# pdf("Figures/CBFGLIS_Pts_Models_Controls_DEGsvsOtherAML_z-scores_1.8.20.pdf", height = 10, width=20)
HA_AML <- ComplexHmap(mat = logCPM[DEGs.AML, heatmap.samps$Sample], 
                      name = "Z-Scores",
                      scale=TRUE,
                      hmap_anno_obj = anno$annoColumn)

HA_AML
# dev.off()
```

```{r}
anno.NBM <- create_HA_Labs_Hmap(expn=logCPM[DEGs.NBM,CBFGLIS_samples$Sample],
                    geneList = DEGs.NBM,
                    CDE = CBFGLIS_samples,
                    cc = cc_heatmaps,
                    cols = Cols)

HA_NBM <- ComplexHmap(mat = logCPM[DEGs.NBM,CBFGLIS_samples$Sample],
                      name = "Z-Scores",
                      scale = TRUE,
                      hmap_anno_obj = anno.NBM$annoColumn)
```

```{r fig.height=10, fig.width=20}
# pdf("Figures/CBFGLIS_Pts_Models_Controls_DEGsvsNBM_Heatmap.pdf", height = 10, width=20)
HA_NBM
# dev.off()
```

## UMAP 

```{r}
bulk <- !grepl("CordBlood|CD34_PB", samps_all$Group) & 
  !grepl("AML, NOS", samps_all$AML_Subtype) & 
  !grepl("No.Primary.Fusion", samps_all$AML_Subtype) 

samps_bulk <- samps_all[bulk,] 
samps_bulk$AML_Subtype <- ifelse(grepl("KMT2A|CBFB|RUNX1|CBFA2T3-GLIS2|NBM", samps_bulk$AML_Subtype),samps_bulk$AML_Subtype, "Other")

table(samps_bulk$AML_Subtype)


newData <- samps_all[grepl("CordBlood", samps_all$Group),] 
table(newData$Comparisons)
```

```{r}
Cols <- c("AML_Subtype","Group",
          "Batch","Time_Point_Group","Tissue")

all(Cols %in% colnames(samps_bulk))

colors37 = c("#466791","#60bf37","#953ada","#4fbe6c","#ce49d3","#a7b43d","#5a51dc","#d49f36","#552095","#507f2d","#db37aa","#84b67c","#a06fda","#df462a","#5b83db","#c76c2d","#4f49a3","#82702d","#dd6bbb","#334c22","#d83979","#55baad","#dc4555","#62aad3","#8c3025","#417d61","#862977","#bba672","#403367","#da8a6d","#a79cd4","#71482c","#c689d0","#6b2940","#d593a7","#895c8b","#bd5975")

cc <- colorCodes_aheatmap(df=samps_bulk[,Cols])
cc$AML_Subtype["Other"] <- "bisque"
cc$AML_Subtype["NBM"] <- "grey70"
cc$AML_Subtype["CBFB-MYH11"] <- "skyblue2"
cc$AML_Subtype["KMT2A"] <- "khaki3"
cc$AML_Subtype["RUNX1-RUNX1T1"] <- "burlywood3"
cc$AML_Subtype["CBFA2T3-GLIS2"] <- "mediumorchid2"


par(mar=c(10,5,5,5))
barplot(rep(1,length(cc$AML_Subtype)),col = cc$AML_Subtype, names.arg = names(cc$AML_Subtype), las=2)
```

```{r}
suppressPackageStartupMessages(library(DelayedArray))

#TFIDF TRANSFORMED Counts
# Term Frequency - Inverse Document Frequency (TF-IDF) 
# https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6101073/

cts_subset <- cts_all[,c(samps_bulk$Sample, newData$Sample)]
cell_total <- apply(cts_subset, 2, sum)
geomean <- exp(mean(log(cell_total)))
sf <- cell_total/geomean
sf.scaled <- t(t(cts_subset)/sf)


tf_all <- seqGlue::tf_idf_transform(sf.scaled)
# tf_alllog <- tf_idf_transform(log2(sf.scaled+1))

gc()

# Mean vs Dispersion Feature Selection 
obj <- seqGlue::calc_dispersion(as.matrix(cts_subset), removeOutliers = TRUE) #removes outlier genes/transcripts based on cooks distance
gc()

sg_all <- seqGlue::get_selected_genes(seqGlue::select_genes(obj, top_n=NULL))
tf_all <- tf_all[sg_all,]
# tf_alllog <- tf_alllog[sg_all,]

dim(tf_all) 
# dim(tf_alllog) 
range(tf_all)


##### PCA (doesnt add enough clarity to makeu p for the runtime)
# norm_cts_log <- log2(sf.scaled+1)
# norm_cts_scaled <- t(scale(t(norm_cts_log), center = TRUE, scale = TRUE))
# # pca_bulk <- run_jackstraw(TFIDF_Matrix = norm_cts_scaled[sg_all,])
# # pca_bulk <- readRDS("Jackstraw_pca_bulk_samples_2021-05-05.RDS")
# 
# # length(pca_bulk$input_features)
# # pca_bulk$N_comp #72
# # saveRDS(pca_bulk, "Jackstraw_pca_bulk_samples_2021-05-05.RDS")
```


n_neighbors	
The size of local neighborhood (in terms of number of neighboring sample points) used for manifold approximation. Larger values result in more global views of the manifold, while smaller values result in more local data being preserved.

min_dist	
The effective minimum distance between embedded points. Smaller values will result in a more clustered/clumped embedding where nearby points on the manifold are drawn closer together, while larger values will result on a more even dispersal of points. 



```{r}
dds <- DESeq2::DESeqDataSetFromMatrix(round(cts_subset, digits = 0),
                                      colData = rbind(samps_bulk, newData),
                                      design = ~ 1)


vst <- DESeq2::vst(dds, blind = TRUE)
dim(vst)


in_vst <- SummarizedExperiment::assay(vst)[sg_all,]

dim(in_vst)
head(in_vst[,1:5])
```

```{r}
umap_bulk <- UMAP_workflow(TFIDF_Matrix = in_vst, 
                           scale_data=FALSE,
                           input_features = sg_all,
                            samples_vector = samps_bulk$Sample,
                            sample_info_df = samps_bulk,
                            Columns_for_Plots = Cols,
                            cc = cc, 
                            min_dist = 0.01,
                            n_neighbors=35,
                            k2=10,
                            res2=0.025)



# saveRDS(umap_bulk,"TARGET_AML_sg2000_md0.005_nn40_UMAP_primary.fusion_bulk_samples_05.12.21.RDS")
 # saveRDS(umap_bulk,"TARGET_AML_sg6678_md0.01_nn38_UMAP_primary.fusion_bulk_samples_05.12.21.RDS")

# saveRDS(umap_bulk,"TARGET_AML_sg6678_md0.01_nn35_UMAP_primary.fusion_bulk_samples_05.12.21.RDS")
```


Best Params with TFIDF
___________
nn = 5-7 
min_dist = 0.05 - 0.005


Best Params with VST
______________
all samps
nn=12
min_dist = 0.005

Has Known Fusion samps
nn=40
min_dist=0.005

```{r fig.height=10, fig.width=10}
umap_bulk$umap_2D_scatter[[1]]
```

```{r}
scatter_plots_3d(umap_workflow_res = umap_bulk, 
                 Group_Column = "AML_Subtype", 
                 Cols=c("Sample",Cols), 
                 cc=cc$AML_Subtype)
```

```{r fig.height=10, fig.width=10}
# umap_bulk$cluster_plots1$scatter
# umap_bulk$cluster_plots1$barplot
```

### Add in the Models to the UMAP Embedding

```{r}
umap_models <- uwot::umap_transform(t(as.matrix(in_vst[,newData$Sample])),
                                    umap_bulk$umap_model,
                                    verbose = TRUE)

dim(umap_models)
```

```{r}
umap_models.df <- read.csv("UMAP/TARGET_AML_sg6678_md0.01_nn35_UMAP_primary.fusion_bulk_CBFLGISmodel_samples_05.12.21.csv")

# umap_models.df <- bind_cols(data.frame(umap_models, fix.empty.names=TRUE),
#                             newData) %>% 
#   rename_at(vars(X1:X3), ~c("x","y","z")) %>% 
#   mutate(AML_Subtype=Comparisons) %>% 
#   bind_rows(umap_bulk$umap_res) %>% 
#   mutate(shape=case_when(
#           grepl("Week6", AML_Subtype) ~ 22,
#           grepl("Week12", AML_Subtype) ~ 24,
#           grepl("CordBlood", AML_Subtype) ~ 21,
#           TRUE ~ 19)) %>% 
#   mutate(pt_size=case_when(
#           # grepl("CBFGLIS_CordBlood", AML_Subtype) ~ 3,
#           grepl("CBFA2T3-GLIS2", AML_Subtype) ~ 3,
#           TRUE ~ 1.0)) %>% 
#   mutate(Time_Point_Group=factor(Time_Point_Group, 
#                                  levels=c("Week1","Week6","Week12"))) 
#   # rename_at(vars(x,y,z), ~c("UMAP_1", "UMAP_2", "UMAP_3"))

# head(umap_models.df)
dim(umap_models.df)

# write.csv(umap_models.df, "TARGET_AML_sg6678_md0.01_nn35_UMAP_primary.fusion_bulk_CBFLGISmodel_samples_05.12.21.csv", row.names = FALSE)


# openxlsx::addWorksheet(supp_table,
#                        sheetName = "UMAP Results")
# openxlsx::writeData(supp_table,
#                     sheet = "UMAP Results",
#                     dplyr::select(umap_models.df, x:z, Sample, USI, AML_Subtype,Time_Point_Group,Tissue),
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```


```{r}
cc_newData <- cc
cc_newData$AML_Subtype["EC_Week1_GFP_CordBlood"] <- "springgreen4"
cc_newData$AML_Subtype["EC_Week1_CBFGLIS_CordBlood"] <- "blue2"
cc_newData$AML_Subtype["EC_Week6_CBFGLIS_CordBlood"] <- "blue2"
cc_newData$AML_Subtype["EC_Week12_CBFGLIS_CordBlood"] <- "blue2"

cc_newData$AML_Subtype["Myeloid_Week1_GFP_CordBlood"] <- "orange"
cc_newData$AML_Subtype["Myeloid_Week1_CBFGLIS_CordBlood"] <- "deeppink"
cc_newData$AML_Subtype["Myeloid_Week6_CBFGLIS_CordBlood"] <- "deeppink"



cc_newData[["AML_point_color"]] <- cc_newData$AML_Subtype
cc_newData$AML_point_color[grep("CordBlood", names(cc_newData$AML_point_color))] <- "black"



# cc_newData$AML_Subtype
par(mar=c(10,5,5,5))
barplot(rep(1,length(cc_newData$AML_Subtype)),col = cc_newData$AML_Subtype, names.arg = names(cc_newData$AML_Subtype), las=2)
```


```{r}
umap_bulk_models <- umap_bulk
umap_bulk_models$umap_res <- umap_models.df

d3_models <- scatter_plots_3d(umap_workflow_res = umap_bulk_models, 
                 Group_Column = "AML_Subtype", 
                 Cols=c("Sample",Cols), 
                 cc=cc_newData$AML_Subtype)

d3_models
# htmlwidgets::saveWidget(d3_models,"TARGET_AML_sg6678_md0.01_nn35_UMAP_primary.fusion_bulk_CBFLGISmodel_samples_05.12.21.html", background = "black")
```

```{r}
# table(umap_models.df$AML_Subtype, umap_models.df$pt_size)

# table()
# table(umap_models.df$pt_size,
#       umap_models.df$AML_Subtype)
```

The UMAP will be designed based on your example figure:
bulk samples into these groups:
0. NBM (light grey, < pt size)
1.CBFB-MYH11 (light/pastel colors, < pt size)
2. RUNX1-RUNX1T1 (light/pastel colors, < pt size)
3. KMT2A (light/pastel colors, < pt size)
4. Others  (light/pastel colors, < pt size)
5. CBFGLIS (light purple, > pt size)
*Labels will be placed next to these groups
 
 
Then Models:
EC_Week1_CBFGLIS_CordBlood (no outline, > pt size, blue )
EC_Week6_CBFGLIS_CordBlood  (dashed outline, > pt size, blue )
EC_Week12_CBFGLIS_CordBlood      (solid outine, > pt size, blue)
Myeloid_Week1_CBFGLIS_CordBlood (no outline, > pt size, red )
Myeloid_Week7_CBFGLIS_CordBlood    (dashed outline, > pt size, red)
EC_Week1_GFP_CordBlood (no outline, < pt size, dark  green)
MC_Week1_GFP_CordBlood (no outline, < pt size, green)
 
 

```{r fig.height=6, fig.width=20}
dims <- list(c("x","y"), c("x","z"), c("y", "z"))
plots <- lapply(dims, make_umap_2d,umap_res.df=umap_models.df, colors_list=cc_newData)

# pdf("UMAP/TARGET_AML_sg6678_md0.01_nn35_UMAP_primary.fusion_bulk_CBFLGISmodel_samples_06.11.21.pdf", height = 6, width = 20)
grid.arrange(grobs=plots, ncol=3, widths=c(0.46, 0.31, 0.32))
# dev.off()
```


### 3D Static
 
```{r}
# devtools::install_github("AckerDWM/gg3D")
library(gg3D)
library(cowplot)
# library(svglite)
```

```{r fig.height=7, fig.width=10}
# NOTE: Cannot do the standard 'hacky' way of subsetting your ggplot data and adding the group of interest a second layer, such as to make the points bigger or add stroke etc. 
# 
# Instead, use the whole dataset as input for a second layer for a group of interest. Change all unwanted color/fill values to NA. Then set the scale_XXX_disrete/scale_XXXX_manual with na.value=NA and/or with na.translate=FALSE to be transparent. Othersize the axes/rotation do not match up because a new min/max for x,y,z coors are used to draw the axes. 

# theta=-120
# phi=-10

ggplot(umap_models.df, 
       aes(x=x, y=y, z=z, fill=AML_Subtype)) + 
  theme_void() +
  axes_3D() +
  stat_3D(geom="point", shape=21, size=umap_models.df$pt_size, phi=45, theta=15) +
  scale_fill_manual(values=cc_newData$AML_Subtype)

```



## Example

```{r}
library(snedata)
# snedata::download_mnist()
# Add new points to an existing embedding
mnist_train <- head(mnist, 60000)
mnist_test <- tail(mnist, 10000)

# You must set ret_model = TRUE to return extra data we need
# coordinates are in mnist_train_umap$embedding
mnist_train_umap <- umap(mnist_train, verbose = TRUE, ret_model = TRUE)
mnist_test_umap <- umap_transform(mnist_test, mnist_train_umap, verbose = TRUE)
```


#Perform Differential Expression

* pairwise for a few initial comparisions
* CBFGLIS binding sites from ChIP-seq from Thomas Mercher




### EC Co-Culture

```{r}
EC.comparisons <-  CBFGLIS_samples %>% #CBFGLIS.models %>% 
  filter(!grepl("sort", Sample), !grepl("Week9", Time_Point_Group)) %>% 
  filter(grepl("EC", Culture_condition)) %>% 
  filter(grepl("GFP|CBFA2T3", Transduction_condition)) %>%
  mutate(EC_Groups_GeneTargets=case_when(
      grepl("EC_Week1_GFP", Comparisons) ~ "GFP", 
      grepl("Week(6|12)_CBFGLIS", Comparisons) ~ "CBFGLIS")) %>% 
  mutate(USI=Sample) %>% 
  set_rownames(.$Sample)

# table(EC.comparisons$Transduction_condition, EC.comparisons$Time_Point_Group)
table(EC.comparisons$Comparisons)
# table(EC.comparisons$EC_Groups_GeneTargets)
# head(EC.comparisons[,1:5])
```

```{r}
ids2symbols <- geneIDs %>% 
  dplyr::select(gene_name, gene_id)

dim(ids2symbols) #58263     2
head(ids2symbols)
dim(in_counts) #58263   182
# write.csv(ids2symbols,"References/Gencode_v29_IDs_to_GeneSymbols.csv", row.names = F)
```

```{r}
ref <- "GFP"
conditions <- c("Week1","Week6", "Week12")
cols.colorbar <- c("AML_Subtype", "Transduction_condition", "Culture_condition", "Time_Point_Group")

DEGs.EC <- lapply(conditions,function(condition){
  df <- filter(EC.comparisons, grepl(ref,Transduction_condition) | 
         (grepl("GLIS",Transduction_condition) & Time_Point_Group == condition)) %>% 
    set_rownames(.$Sample)
  
  DE <- twoGroups_DEGs(expnData = in_counts,
                       clinData = df,
                       col="Transduction_condition",
                       ref = ref,
                       ids2symbols = "References/Gencode_v29_IDs_to_GeneSymbols.csv",
                       anno = FALSE,
                       SkipPlots = TRUE,
                       Custom.Cols = cols.colorbar)
  
})

names(DEGs.EC) <- conditions
# saveRDS(DEGs.EC, "DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_EC.culture_DEGs_5.21.21.RDS")
```

```{r}
DEGs.EC <- readRDS("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_EC.culture_DEGs_5.21.21.RDS")
# DEGs.EC$Week1$phenovector
# dim(DEGs.EC$Week1$DE$Voom$E) #19492     6
# head(DEGs.EC$Week1$DE$eBayesFit$p.value[order(DEGs.EC$Week1$DE$eBayesFit$p.value)])

# dir.create("DEGs/for_volcano_plots/")
raw_res <- lapply(names(DEGs.EC[2:3]), function(x){
  fit <- DEGs.EC[[x]]$DE$eBayesFit
  df <- data.frame(gene=rownames(fit[["coefficients"]]),
              log2FC=fit[["coefficients"]][,1],
              p.value=fit[["p.value"]][,1]) %>% 
    mutate(FDR=p.adjust(p.value, method = "BH")) %>% 
    left_join(., select(geneIDmap, gene_name, gene_id),
              by=c("gene"="gene_name")) %>% 
    arrange(desc(log2FC), desc(FDR)) %>% 
    select(gene, gene_id, everything())
  
  # filename <- paste0("DEGs/for_volcano_plots/TARGET_AML_CBFGLIS.model_", x, "_vs_GFP_EC.culture_DEGs_5.21.21.csv")
  # con <- file(filename, open="wt")
  # writeLines(paste("# Differentially expressed genes in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), con)
  # writeLines(paste("# Culture Condition: EC co-culture"), con)
  # writeLines(paste("# Input Samples:", x, "CBFGLIS_Cordblood (N=4) vs  GFP_Cordblood Week1 (N=3)"), con)
  # write.csv(df, con,row.names = F)
  # close(con)
})

```

```{r}
# table(DEGs.EC$Week6$phenovector)
DEGs.EC.week7 <- read.csv("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week7_EC.culture_DEGs_5.21.21.csv", comment.char = "#")

# DEGs.EC.week7 <- extract_DEGs(DEGs.EC$Week6) %>% 
#   left_join(., ids2symbols, by=c("gene"="gene_name")) 

head(DEGs.EC.week7)
tail(DEGs.EC.week7)
dim(DEGs.EC.week7) #4076    9

# DEGs.EC.week7 %>% 
#   filter(grepl("FOLR1", gene))


# conWeek7 <- file("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week7_EC.culture_DEGs_5.21.21.csv", open="wt")
# writeLines(paste("# Differentially expressed genes in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), conWeek7)
# writeLines(paste("# Culture Condition: EC co-culture"), conWeek7)
# writeLines(paste("# Input Samples: Week7 CBFGLIS_Cordblood (N=4) vs  GFP_Cordblood Week1 (N=3)"), conWeek7)
# write.csv(DEGs.EC.week7, conWeek7, row.names = F)
# close(conWeek7)


# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFGLIS HSPCs DEGs EC week6")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFGLIS HSPCs DEGs EC week6",
#                     data.frame(), #blank so I can simply add in the table with the comments
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```

```{r}
DEGs.EC.week12 <- read.csv("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week12_EC.culture_DEGs_5.21.21.csv",comment.char = "#")
# DEGs.EC.week12 <- extract_DEGs(DEGs.EC$Week12) %>% 
#   left_join(., ids2symbols, by=c("gene"="gene_name"))
  # filter(grepl("^ENSG", gene_id))

head(DEGs.EC.week12)
dim(DEGs.EC.week12) #4083    9


# DEGs.EC.week12 %>% 
#   filter(grepl("FOLR1", gene))


# conWeek12 <- file("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week12_EC.culture_DEGs_5.21.21.csv", open="wt")
# writeLines(paste("# Differentially expressed genes in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), conWeek12)
# writeLines(paste("# Culture Condition: EC co-culture"), conWeek12)
# writeLines(paste("# Input Samples: Week12 CBFGLIS_Cordblood (N=4) vs  Week1 GFP_Cordblood (N=3)"), conWeek12)
# write.csv(DEGs.EC.week12, conWeek12, row.names = F)
# close(conWeek12)



# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFGLIS HSPCs DEGs EC week12")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFGLIS HSPCs DEGs EC week12",
#                     data.frame(), #blank so I can simply add in the table with the comments
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```


#### For Target genes 

```{r}
ref <- "GFP"
conditions <- c("Week1","Week6", "Week12")
cols.colorbar <- c("AML_Subtype", "Transduction_condition", "Culture_condition", "Time_Point_Group")

forTargets <- twoGroups_DEGs(expnData = in_counts, 
                       clinData = EC.comparisons[!is.na(EC.comparisons$EC_Groups_GeneTargets),],
                       col="EC_Groups_GeneTargets", 
                       ref = ref, 
                       ids2symbols = "References/Gencode_v29_IDs_to_GeneSymbols.csv",
                       anno = FALSE,
                       SkipPlots = TRUE,
                       Custom.Cols = cols.colorbar)

dim(forTargets$DE$Voom$E) #19843    11
# saveRDS(forTargets, "TARGET_AML_CBFGLIS.models_Week7and12_vs_GFP_week1_DEGs.RDS")
```

```{r}
DE.forTargets <- extract_DEGs(forTargets) 
  #   mutate(SialicAcidPathway=case_when(gene %in% sialicAcid$gene_name.stable.ID ~ "Yes")) %>%
  #   mutate(Cell_Adhesion_Gene=ifelse(gene %in% unlist(GO.Adhesion), "Yes", NA)) %>%
  #   mutate(CancerTestesAntigen_CTA=case_when(
  #       gene %in% CTAs$Family.member ~ "CTA",
  #   TRUE  ~ "")) %>% 
  # dplyr::select(gene,SialicAcidPathway:CancerTestesAntigen_CTA,Treatment.type,everything())


# con <- file("TARGET_AML_CBFGLIS.models_Week7and12_vs_GFP_week1_DEGs.csv", open="wt")
# writeLines(paste("# Differentially expressed genes in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), con)
# writeLines(paste("# Culture Condition: EC co-culture"), con)
# writeLines(paste("# Input Samples: Week7 and Week12 CBFGLIS_Cordblood (N=8) vs  GFP_Cordblood (N=3)"), con)
# write.csv(DE.forTargets, con, row.names = FALSE)
# close(con)

head(DE.forTargets)
# dim(DE.forTargets) #4048    8
```

```{r}
forTargets$DE$Voom$E["FOLR1",]

forTargets$DE$eBayesFit$coefficients["FOLR1",]
FDR <- p.adjust(forTargets$DE$eBayesFit$p.value, method = "BH")
names(FDR) <- rownames(forTargets$DE$eBayesFit$p.value)

FDR["FOLR1"]
```


## Myeloid Culture

```{r}
MC.comparisons <- CBFGLIS_samples %>% 
  filter(!grepl("scbulkleftovers|pool", Sample)) %>% 
  filter(grepl("Myeloid", Culture_condition)) %>% 
  filter(grepl("GFP|CBFA2T3", Transduction_condition)) %>% 
  filter(!c(grepl("GFP", Transduction_condition) & grepl("Week12|Week6", Time_Point_Group))) %>% 
  mutate(USI=Sample) %>% 
  set_rownames(.$Sample)

table(MC.comparisons$Transduction_condition, MC.comparisons$Time_Point_Group)

table(MC.comparisons$Comparisons)
# View(MC.comparisons)
```

```{r}
ref <- "GFP"
conditions <- c("Week1","Week6", "Week12")
cols.colorbar <- c("AML_Subtype", "Transduction_condition", "Culture_condition", "Time_Point_Group")
```

```{r}
DEGs.MC <- lapply(conditions,function(condition){
  df <- filter(MC.comparisons, grepl(ref,Transduction_condition) | 
         (grepl("GLIS",Transduction_condition) & Time_Point_Group == condition)) %>% 
    set_rownames(.$Sample)
  
  DE <- twoGroups_DEGs(expnData = in_counts,
                       clinData = df,
                       col="Transduction_condition",
                       ref = ref,
                       ids2symbols = "References/Gencode_v29_IDs_to_GeneSymbols.csv",
                       anno = FALSE,
                       SkipPlots = TRUE,
                       Custom.Cols = cols.colorbar)

})

names(DEGs.MC) <- conditions
# saveRDS(DEGs.MC, "DEGs/00_Archive/TARGET_AML_CBFGLIS.model_vs_GFP_MC.culture_DEGs.RDS")
```

```{r}
DEGs.MC <- readRDS("DEGs/00_Archive/TARGET_AML_CBFGLIS.model_vs_GFP_MC.culture_DEGs.RDS")
DEGs.MC$Week1$phenovector
dim(DEGs.MC$Week1$DE$Voom$E) #18603     6
# head(DEGs.MC$Week1$DE$eBayesFit$p.value[])
```

```{r}
DEGs.MC.week7  <- read.csv("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week7_MC.culture_DEGs_5.21.21.csv", comment.char = "#")

# dim(DEGs.MC$Week7$DE$Voom$E) 18830     6
# DEGs.MC.week7 <- extract_DEGs(DEGs.MC$Week6) %>% 
#   left_join(., ids2symbols, by=c("gene"="gene_name"))
  # filter(grepl("^ENSG", gene_id))

head(DEGs.MC.week7)
dim(DEGs.MC.week7) #3797    8



# conWeek7MC <- file("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week7_MC.culture_DEGs_5.21.21.csv", open="wt")
# writeLines(paste("# Differentially expressed genes in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), conWeek7MC)
# writeLines(paste("# Culture Condition: Myeloid culture"), conWeek7MC)
# writeLines(paste("# Input Samples: Week7 CBFGLIS_Cordblood (N=3) vs  Week1 GFP_Cordblood (N=3)"), conWeek7MC)
# write.csv(DEGs.MC.week7, conWeek7MC, row.names = F)
# close(conWeek7MC)



# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFGLIS HSPCs DEGs MC week6")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFGLIS HSPCs DEGs MC week6",
#                     data.frame(), #blank so I can simply add in the table with the comments
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```

```{r}
DEGs.MC.week12 <- read.csv("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week12_MC.culture_DEGs_5.21.21.csv", comment.char = "#")

# DEGs.MC.week12 <- extract_DEGs(DEGs.MC$Week12) %>% 
#   left_join(., ids2symbols, by=c("gene"="gene_name"))

head(DEGs.MC.week12)
dim(DEGs.MC.week12) #2055    8


# conWeek12MC <- file("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week12_MC.culture_DEGs_5.21.21.csv", open="wt")
# writeLines(paste("# Differentially expressed genes in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), conWeek12MC)
# writeLines(paste("# Culture Condition: Myeloid culture"), conWeek12MC)
# writeLines(paste("# Input Samples: Week12 CBFGLIS_Cordblood (N=3) vs  Week1 GFP_Cordblood (N=3)"), conWeek12MC)
# write.csv(DEGs.MC.week12, conWeek12MC, row.names = F)
# close(conWeek12MC)




# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFGLIS HSPCs DEGs MC week12")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFGLIS HSPCs DEGs MC week12",
#                     data.frame(), #blank so I can simply add in the table with the comments
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```


# DEGs in Models and Patients

```{r}
# CBFGLISvsAML$gene_name
```

```{r}
DEGs.list <- lapply(c(DEGs.EC[-1], DEGs.MC[-1]), function(x) extract_DEGs(x)) 
names(DEGs.list) <- paste0(rep(c("EC_","MC_"), each=2), names(DEGs.list))

length(DEGs.list)
names(DEGs.list)

n.common <- sapply(DEGs.list, function(x) length(intersect(pull(x, gene), CBFGLISvsAML$gene_name)))
```

```{r}
degs.common <- data.frame(Culture_Condition=rep(c("EC","MC"), each=3),
                          Time_Point=rep(paste("Week",c(1,7,12)))) %>% 
  mutate(Comparison=rep(paste(rep("GFP_CordBlood", 3), "x", 
                              paste("CBFGLIS_CordBlood")),2), 
         Number_DEGs=c(0,nrow(DEGs.list$EC_Week7), nrow(DEGs.list$EC_Week12), 
                       0, nrow(DEGs.list$MC_Week7), nrow(DEGs.list$MC_Week12)), 
         Number_Shared_DEGs=c(0, n.common[1:2], 
                              0, n.common[3:4]), 
         Percent_Shared_DEGs=round((Number_Shared_DEGs/Number_DEGs)*100, digits = 2))


# degs.common
# write.csv(degs.common, "TARGET_AML_CBFGLIS.models_DEGs_Summary_Table_5.21.21.csv", row.names = F)
```





# Expression plots Genes of Interest

```{r}
library(ggbeeswarm)
library(patchwork)
```

```{r}
simple_boxplot <- function(df){
  box <- ggplot(df, 
                  aes(x=Comparisons, y=log2CPM, fill=Comparisons, color=Comparisons)) +
  geom_jitter() +
  geom_boxplot(alpha=0.6, color="black") +
  facet_wrap(~gene) +
  scale_fill_brewer(palette = "Paired") +
    scale_color_brewer(palette = "Paired") +
  labs(x="",y="Expression (log2 CPM)") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=25, vjust=1, hjust=1, size=10),
        panel.border = element_rect(color="black", fill=NA),
        plot.margin = margin(l=15, unit="mm"),
        strip.text = element_text(size=14),
        legend.text = element_text(size=14))
  
  return(box)
}
```

```{r}
# filter(geneIDmap, grepl("^KIT$|^ERG$|GATA1$", gene_name))
```

```{r}
samps_for_violin <- CBFGLIS_samples %>% 
  filter(!grepl("pool|sort|pos|D63.EC", Sample),
         !c(grepl("replicate", Sample) & Group=="NBM"),
         !grepl("mock",Sample)) %>% 
  filter(!grepl("(Week6|Week12)_GFP|CD34_PB", Comparisons)) %>% 
  mutate_at(vars(Comparisons), ~factor(.,  levels=c(
                                                    # "CD34_PB",
                                                    "EC_Week1_GFP_CordBlood",
                                                    "EC_Week1_CBFGLIS_CordBlood",
                                                    "EC_Week6_CBFGLIS_CordBlood",
                                                    "EC_Week12_CBFGLIS_CordBlood", 
                                                    "Myeloid_Week1_GFP_CordBlood", 
                                                    "Myeloid_Week1_CBFGLIS_CordBlood", 
                                                    "Myeloid_Week6_CBFGLIS_CordBlood",
                                                    "Myeloid_Week12_CBFGLIS_CordBlood",
                                                    "CBFA2T3-GLIS2",
                                                    "NBM"))) %>% 
  filter(!grepl("Myeloid_Week12_CBFGLIS_CordBlood", Comparisons)) %>% 
  droplevels() %>% 
  mutate(Facet=factor(ifelse(grepl("CD34_PB|NBM|CBFA2T3-GLIS2",Comparisons), "Primary Samples", "CBFA2T3-GLIS2 HSPCs"), 
                        levels=c("Primary Samples","CBFA2T3-GLIS2 HSPCs")))
  

table(samps_for_violin$Comparisons, useNA='ifany')
table(samps_for_violin$Facet, useNA='ifany')
# table(samps_for_violin$Culture_condition, useNA='ifany')
# unique(samps_for_violin$Comparisons[order(samps_for_violin$Comparisons)])
```

```{r}
genes.of.interest <- CPM[grep("^ERG$|GATA1$|^BMP2$|^BMP4$|^WT1$", rownames(CPM)), samps_for_violin$Sample] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  gather(Sample, CPM, -gene) %>% 
  mutate(log2CPM=log2(CPM+1)) %>% 
  left_join(., samps_for_violin, by="Sample")  %>% 
  mutate(gene=factor(gene, levels=c("ERG","BMP2","BMP4","GATA1","WT1"))) %>% 
  mutate(Size=ifelse(Facet=="Primary Samples", 0.5, 2.0)) %>% 
  arrange(gene)


head(genes.of.interest)
# dim(genes.of.interest)
# table(genes.of.interest$Comparisons, useNA='ifany')
# table(genes.of.interest$Size, genes.of.interest$Comparisons)


# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFA2T3-GLIS2 Associated Genes")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFA2T3-GLIS2 Associated Genes",
#                     dplyr::select(genes.of.interest, Sample, USI, gene, CPM,
#                                   AML_Subtype=Comparisons,Time_Point_Group,Tissue),
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```

```{r}
cc.cultures <- c(
                 "EC_Week1_GFP_CordBlood"="springgreen4",
                 "EC_Week1_CBFGLIS_CordBlood"="blue2",
                 "EC_Week6_CBFGLIS_CordBlood"="blue2",
                 "EC_Week12_CBFGLIS_CordBlood"="blue2", 
                 "Myeloid_Week1_GFP_CordBlood"="orange2", 
                 "Myeloid_Week1_CBFGLIS_CordBlood"="deeppink", 
                 "Myeloid_Week6_CBFGLIS_CordBlood"="deeppink",
                 "CBFA2T3-GLIS2"="mediumorchid2",
                 "NBM"="black")
```

```{r fig.height=6, fig.width=30}
breaks_list <- list("GATA1"=c(0,15,30,45),
                    "BMP2"=c(0,20,40,60),
                    "BMP4"=c(0,5,10,15),
                    "ERG"=c(0,60,120, 180))
breaks_list <- breaks_list[levels(genes.of.interest$gene)]

boxplots <- lapply(as.character(unique(genes.of.interest$gene)), function(x){

  
  make_facet_boxplots(genes.of.interest = droplevels(genes.of.interest[genes.of.interest$gene==x,]),
                      breaks=breaks_list[[x]])
})

# boxplots

# pdf("Figures/Boxplots/TARGET_AML_CBFGLIS.models_CBFGLIS.patients_GATA1_BMP2_BMP4_ERG_Expression_boxplots.pdf", height=6, width = 30)
gridExtra::grid.arrange(grobs=boxplots, nrow=1)
# dev.off()
```

```{r fig.width=15, fig.height=6}
my_comparisons <- list(c("EC_Week1_GFP_CordBlood", "EC_Week1_CBFGLIS_CordBlood"),
                       c("EC_Week1_GFP_CordBlood", "EC_Week6_CBFGLIS_CordBlood"),
                       c("EC_Week1_GFP_CordBlood", "EC_Week12_CBFGLIS_CordBlood"),
                       c("Myeloid_Week1_GFP_CordBlood", "Myeloid_Week1_CBFGLIS_CordBlood"),
                       c("Myeloid_Week1_GFP_CordBlood", "Myeloid_Week6_CBFGLIS_CordBlood"),
                       c("NBM","CBFA2T3-GLIS2"))


# ggplot(data=genes.of.interest, aes(x=Comparisons, y=CPM, fill=Comparisons)) +
#       geom_boxplot(alpha=0.2,outlier.shape = NA,  inherit.aes = TRUE)  +
#   stat_compare_means(comparisons = my_comparisons, method="wilcox.test", 
#                      method.args = list(alternative="two.sided"), 
#                      label="p.signif") + #"p.signif"
#   facet_wrap(~gene) +
#   theme(axis.text.x = element_text(angle=25, vjust=1, hjust=1))
```




```{r}
DEGs.list.objs <- ls(pattern = "DEGs.EC|DEGs.MC")
DEGs.list <- lapply(DEGs.list.objs, function(x) get(x))
names(DEGs.list) <- DEGs.list.objs

# sapply(DEGs.list, dim)

DEGs.GOI.df <- lapply(names(DEGs.list), function(x) filter(DEGs.list[[x]], grepl("^ERG$|GATA1$|^BMP2$", gene)) %>% 
         dplyr::select(gene, logFC, adj.P.Val) %>% 
         mutate(Culture=str_split_fixed(x, pattern="\\.", n=3)[,2],
                Time_Point=gsub("7", "6", str_split_fixed(x, pattern="\\.", n=3)[,3]),
                Comparison="vs GFP Week1")) %>% 
  bind_rows() %>% 
  mutate(adj.P.Val=round(adj.P.Val, digits = 3)) %>% 
  mutate(Signif_level=case_when(
                  adj.P.Val > 0.05 ~ "n.s.",
                  adj.P.Val <= 0.0001 ~ "****",
                  adj.P.Val <= 0.001 ~ "***",
                  adj.P.Val <= 0.01 ~ "**",
                  adj.P.Val < 0.05 ~ "*",
                )) %>% 
  arrange(gene, desc(Time_Point))

DEGs.GOI.df
# write.csv(DEGs.GOI.df, "DEGs/TARGET_AML_GenesOfInterest_DEGs.csv", row.names = FALSE)
```


```{r fig.width=10}
EC.box.WT1 <- simple_boxplot(df=filter(genes.of.interest,grepl("WT1", gene))) +
  theme(plot.margin = margin(l=3, unit="cm")) +
  stat_compare_means(comparisons = my_comparisons, method.args = list(alternative="less"), label="p.format")

EC.box.WT1
# ggsave(plot = EC.box.WT1, filename = "Figures/TARGET_AML_CBFGLIS.models_and_Patients_and_NBM_WT1_Expression.pdf", device = "pdf", height = 7, width = 10)
```


```{r fig.height=5, fig.width=15}
EC.box1 <- simple_boxplot(df=filter(genes.of.interest,grepl("GLIS2|EC", Culture_condition)))
EC.box2 <- simple_boxplot(df=filter(genes.of.interest,grepl("GLIS2|EC|NBM|PB", Culture_condition)))
EC.box3 <- simple_boxplot(df=filter(genes.of.interest,grepl("GLIS2|EC", Culture_condition),
                                    !grepl("Week1_CBFGLIS", Comparisons)))

# pdf("Figures/TARGET_AML_CBFGLISpts_CBFGLIS.models_EC_Boxplot.pdf", height = 5, width = 15)
# EC.box1
# dev.off()


# pdf("Figures/TARGET_AML_CBFGLISpts_CBFGLIS.models_NBM_EC_Boxplot.pdf", height = 5, width = 15)
# EC.box2
# dev.off()
# 
# 
# pdf("Figures/TARGET_AML_CBFGLISpts_CBFGLIS.models_LateCultures_EC_Boxplot.pdf", height = 5, width = 15)
# EC.box3
# dev.off()

```

```{r fig.height=5, fig.width=15}
MC.box1 <- simple_boxplot(df=filter(genes.of.interest,grepl("GLIS2|My", Culture_condition)))
MC.box2 <- simple_boxplot(df=filter(genes.of.interest,grepl("GLIS2|My|NBM|PB", Culture_condition)))
MC.box3 <- simple_boxplot(df=filter(genes.of.interest,grepl("GLIS2|My", Culture_condition), 
                                    !grepl("Week1_CBFGLIS", Comparisons)))

# pdf("Figures/TARGET_AML_CBFGLISpts_CBFGLIS.models_MC_Boxplot.pdf", height = 5, width = 15)
# MC.box1
# dev.off()

# pdf("Figures/TARGET_AML_CBFGLISpts_CBFGLIS.models_NBM_MC_Boxplot.pdf", height = 5, width = 15)
# MC.box2
# dev.off()
# 
# 
# pdf("Figures/TARGET_AML_CBFGLISpts_CBFGLIS.models_LateCultures_MC_Boxplot.pdf", height = 5, width = 15)
# MC.box3
# dev.off()

```





# Comparison of Time Points 

```{r}
models.subset <- bind_rows(EC.comparisons, MC.comparisons) %>% 
  set_rownames(.$Sample) %>% 
  mutate_at(vars(Comparisons), ~factor(., levels=c(  "EC_Week1_GFP_CordBlood",
                                                    "EC_Week1_CBFGLIS_CordBlood",
                                                    "EC_Week6_CBFGLIS_CordBlood",
                                                    "EC_Week12_CBFGLIS_CordBlood",
                                                    "Myeloid_Week1_GFP_CordBlood",
                                                    "Myeloid_Week1_CBFGLIS_CordBlood",
                                                    "Myeloid_Week6_CBFGLIS_CordBlood"))) %>%
  filter(!is.na(Comparisons)) %>%  #these are myeloid week 12
  arrange(Comparisons) %>% 
  set_rownames(.$Sample)
  # arrange("Culture_condition","Transduction_condition","Time_Point_Group")

dim(models.subset)
table(models.subset$Comparisons)

```

```{r}
model_counts <- in_counts[,models.subset$Sample]
keep.mod <- rowSums(cpm(model_counts) >= 1) >= 2
cts.mod <- model_counts[keep.mod, ]

dge.mod <- DGEList(counts=cts.mod)
dge.mod <- calcNormFactors(dge.mod,method = "TMMwsp")

logCPM.mod <- edgeR::cpm(dge.mod,log=TRUE,normalized.lib.sizes=TRUE, prior.count=1)
CPM.mod <- edgeR::cpm(dge.mod,log=FALSE,normalized.lib.sizes=TRUE, prior.count=1)

dim(logCPM.mod) #21626    23
head(logCPM.mod[,1:5])
```

```{r}
genes <- lapply(c(DEGs.EC[-1], DEGs.MC[-c(1,3)]),
                function(x) extract_DEGs(x) %>% filter(abs(logFC) > 2) %>%  pull(gene)) %>% 
  unlist() %>% 
  unique()


table(genes %in% rownames(logCPM.mod))
length(genes) #3,408 genes with FC>4
```


```{r fig.height=7, fig.width=12}
len <- 299
col <- colorRampPalette(c("deepskyblue4", "deepskyblue3", "deepskyblue2", "deepskyblue1","white","red1", "red2", "red3", "red4"))(n=len)

# pdf("TARGET_AML_CBFGLIS.models_vs_GFP.cordblood_DEGs_FC.GT.4_Heatmap_7.23.21.pdf", height = 7, width = 12)
p <- pheatmap::pheatmap(mat= t(logCPM.mod[genes,]),
                     color=col,
                     scale = "column",
                     # breaks = Breaks,
                     cluster_rows=FALSE,
                     cluster_cols = TRUE,
                     clustering_method="ward.D2",
                     clustering_distance_cols="euclidean",
                     annotation_names_col=TRUE,
                     annotation_row = models.subset[,c("Culture_condition",
                                                       "Transduction_condition", 
                                                       "Time_Point_Group")],
                     # kmeans_k=2,
                     annotation_colors=list("Culture_condition"=c("EC"="navy","Myeloid"="blue1"),
                                            "Transduction_condition"=c("GFP"="grey","CBFA2T3_GLIS"="red4"), 
                                            "Time_Point_Group"=c("Week1"="pink1","Week6"="pink3", "Week12"="pink4")),
                     gaps_row=c(14),
                     fontsize = 10,
                     fontsize_row=10,
                     fontsize_col=10,
                     show_colnames=FALSE,
                     show_rownames = F)
# dev.off()
```


```{r}
#https://rdrr.io/cran/pheatmap/src/R/pheatmap.r 
scale_rows = function(x){
    m = apply(x, 1, mean, na.rm = T)
    s = apply(x, 1, sd, na.rm = T)
    return((x - m) / s)
}

scale_mat = function(mat, scale){
    if(!(scale %in% c("none", "row", "column"))){
        stop("scale argument shoud take values: 'none', 'row' or 'column'")
    }
    mat = switch(scale, none = mat, row = scale_rows(mat), column = t(scale_rows(t(mat))))
    return(mat)
}

pheatmap.zscores <- scale_mat(mat=t(logCPM.mod[genes,]), scale="column")
dim(pheatmap.zscores)
# head(pheatmap.zscores[,1:5])

# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFGLIS HSPCs Heatmap Z-scores")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFGLIS HSPCs Heatmap Z-scores",
#                     pheatmap.zscores,
#                     keepNA=FALSE,
#                     rowNames=TRUE)
```


# Comparison to CBFGLIS Patient Samples

```{r}
models.subset <- bind_rows(EC.comparisons, MC.comparisons) %>% 
  set_rownames(.$Sample)

dim(models.subset)
table(models.subset$Comparisons)
```

```{r}
CBFGLIS.subset <- models.subset %>% 
  bind_rows(., filter(CBFGLIS_samples, AML_Subtype=="CBFA2T3-GLIS2")) %>% 
  mutate_all(~gsub("CBFA2T3-GLIS2","CBFA2T3.GLIS2", .)) %>% 
  mutate(Comparisons=paste(Culture_condition, Time_Point_Group, AML_Subtype, sep="_")) %>% 
  mutate(Comparisons=gsub("^(CBFA2T3.GLIS2)_.+", "\\1", Comparisons)) 

dim(CBFGLIS.subset) #63 63
# table(CBFGLIS.subset$Time_Point_Group)
table(CBFGLIS.subset$Comparisons)
```

```{r}
CBFGLIS_counts <- in_counts[,CBFGLIS.subset$Sample]
keep.cbf <- rowSums(CBFGLIS_counts) >= 5
cts.cbf <- CBFGLIS_counts[keep.cbf, ]

dge.cbf <- DGEList(counts=cts.cbf)
dge.cbf <- calcNormFactors(dge.cbf,method = "TMMwsp")

logCPM.cbf <- edgeR::cpm(dge.cbf,log=TRUE,normalized.lib.sizes=TRUE, prior.count=1)
CPM.cbf <- edgeR::cpm(dge.cbf,log=FALSE,normalized.lib.sizes=TRUE, prior.count=1)

dim(logCPM.cbf) #48478    66
head(logCPM.cbf[,1:5])
```

```{r}
#So likely due to the fact that CBFGLISvsAML had a huge referecne group of samples (>1,000) a lot more genes met the typical 1CPM cut-off. But to avoid excluding as many DE genes as possible, I lowered the CPM threshold to just 5 counts per row... Still missing 2 genes... 
table(CBFGLISvsAML$gene_name %in% rownames(logCPM.cbf))

CBFGLISvsAML$gene_name[!CBFGLISvsAML$gene_name %in% rownames(logCPM.cbf)]
```

```{r}
genes <- lapply(c(DEGs.EC[-1], DEGs.MC[-1]), function(x) extract_DEGs(x) %>%  pull(gene)) 
names(genes) <- paste0(rep(c("EC_","MC_"), each=2), names(genes))
genes.all <-  c(list("EC_Week1_GFP"=CBFGLISvsAML$gene_name,"EC_Week1"=CBFGLISvsAML$gene_name), genes[1:2],
                list("MC_Week1_GFP"=CBFGLISvsAML$gene_name,"MC_Week1"=CBFGLISvsAML$gene_name), genes[3:4])
 
sapply(genes.all, length)

in.degs <- lapply(genes.all, function(x) intersect(CBFGLISvsAML$gene_name, x))
in.degs <- lapply(in.degs, function(x) intersect(rownames(logCPM.cbf), x))
names(in.degs) <- names(genes.all)
sapply(in.degs, length)

# length(in.degs) #8 
```

```{r}
grps <- unique(CBFGLIS.subset$Comparisons)
# length(grps) #9
# cbind(names(in.degs), grps[-9]) #Matches Order
```

```{r}
scatter_plots <- lapply(1:8, function(i){
  
    comp1 <- CBFGLIS.subset %>% 
        filter(grepl("CBFA2T3.GLIS2", AML_Subtype) | grepl(grps[i],Comparisons))
    degs <- in.degs[[i]]
    cols <- c("CBFA2T3.GLIS2",grps[i])
    print(cols)
    
    plot <- scatter_comparison(df=comp1, cols=cols, genes=degs)
    return(plot)
})
```

```{r fig.height=15, fig.width=15}
grid.arrange(grobs=scatter_plots, ncol=2)
```




# GSEA


### GAGE Resutls

```{r}
library(gageData)
library(gage)

data(egSymb)
data(sigmet.idx.hs)
data(kegg.sets.hs)
kegg.all <- kegg.sets.hs[sigmet.idx.hs]
kegg.all.sym <- lapply(kegg.all, eg2sym)

length(kegg.all.sym)
grep("hedge|wnt|tgf", ignore.case = T,names(kegg.all.sym), value=TRUE)
```

```{r}
gage_res_EC <- lapply(DEGs.EC[c("Week6","Week12")], gage_from_pipeline, type="expn",geneset=NULL)
names(gage_res_EC) <- c("Week6","Week12")
```

```{r}
dim(gage_res_EC$Week6$SigPaths.Up)
dim(gage_res_EC$Week6$SigPaths.Dn)
```

```{r}
save_res <- lapply(names(gage_res_EC), function(x){
  dat <- gage_res_EC[[x]]
  
  for (direction in c("SigPaths.Up", "SigPaths.Dn")){
      fname <- paste("GSEA/TARGET_AML_CBFGLIS.model_vs_GFP_EC", x, direction,"gage_GSEA.csv", sep="_")
      print(fname)
     
      gsea_paths <-  dat[[direction]] %>% 
        as.data.frame() %>% 
        rownames_to_column("pathway") 
      
      con <- file(fname, open="wt")
      writeLines(paste("# Differentially expressed pathways in CBFGLIS transduced cordblood compared to GFP transduced cordblood."), con)
      writeLines(paste("# Culture Condition: EC co-culture"), con)
      writeLines(paste("# Input Samples:", x, "CBFGLIS_Cordblood (N=4) vs  GFP_Cordblood Week1 (N=3)\n"), con)
      
      write.csv(gsea_paths, con, row.names = FALSE)
      close(con)
  }
  
})
```





### PathfindR Results

```{r}
library(pathfindR)
```

```{r}
# res_for_gsea <- lapply(c(DEGs.EC[-1], DEGs.MC[-1]), function(x) extract_DEGs(x) %>%  
#                          dplyr::select(gene, logFC, adj.P.Val)) 
res_for_gsea <- lapply(ls(pattern="DEGs.[EM]"), function(x) get(x) %>%  
                         dplyr::select(gene, logFC, adj.P.Val)) 

# names(res_for_gsea) <- paste0(rep(c("EC_","MC_"), each=2), names(res_for_gsea))
names(res_for_gsea) <- ls(pattern="DEGs.[EM]")
names(res_for_gsea)
```

Error in grid.Call(C_stringMetric, as.graphicsAnnot(x$label)) : 
  rsession: unable to read font `helvetica' @ error/annotate.c/RenderFreetype/1384
  
```{r}
gsea <- lapply(1:length(res_for_gsea), function(i){
  res <- run_pathfindR(input=res_for_gsea[[i]], 
              gene_sets = "KEGG", 
              min_gset_size = 15,
              max_gset_size = 400,
              visualize_enriched_terms=FALSE)
  
  # saveRDS(res, paste0("CBFGLIS.model_", names(res_for_gsea)[i], "_vs_GFP.cordblood_GSEA.RDS"))
})

names(gsea) <- names(res_for_gsea)
```

The p-value is derived from a re-sampling/permutation method, thus the lowest_p and highest_p
so that will change from run-to-run, so not the most ideal for getting identical results/reproducibility.

```{r}
head(gsea$DEGs.EC.week7)
# dim(gsea$DEGs.EC.week7) #146   8
# setdiff(gsea$DEGs.EC.week7$ID, res.rdata$EC_Week7$ID)
# setdiff(res.rdata$EC_Week7$ID, gsea$DEGs.EC.week7$ID)
```

```{r}
head(res.rdata$EC_Week7)
# dim(res.rdata$EC_Week7) #151   8

res.rdata$EC_Week7 %>% 
  filter(ID %in% setdiff(ID, gsea$DEGs.EC.week7$ID))
```

```{r}
res.paths.rdata <- dir(path = "GSEA",pattern="*.RDS", full.names = TRUE)
# res.paths.rdata 

res.rdata <- lapply(res.paths.rdata, readRDS)
names(res.rdata) <- gsub("^.+([EM].+[72])_.+", "\\1", res.paths.rdata)

res.rdata.fdr <- lapply(res.rdata, function(x){
  x %>% 
    mutate(FDR=p.adjust(lowest_p, method = "BH")) %>% 
    select(ID:Term_Description, FDR, everything())
})


# lapply(res.rdata.fdr, head)
# lapply(res.rdata.fdr, dim)
```

```{r}
# saved <- lapply(names(res.rdata), function(x){
#  file <- gsub(".RDS",".csv", res.paths.rdata[grep(x, res.paths.rdata)])
#  write.csv(res.rdata[[x]], file = file, row.names = FALSE)
# })
```

```{r}
paths_of_interest <- lapply(names(res.rdata), function(x){
  
  comp <- gsub("7","6",x) %>% 
    str_split(.,"_") %>% 
    unlist() 
  
  comp <- paste0(comp[1], "_CBFGLIS_", comp[2],"_vs_GFP_Week1")

  res.rdata[[x]] %>%
    filter(grepl("Hedge|TGF-beta|Wnt", Term_Description, ignore.case = TRUE)) %>%
    mutate(Comparison=comp) %>%
    select(Comparison, everything())
  
}) %>%
  bind_rows()

paths_of_interest

# write.csv(paths_of_interest, "GSEA/TARGET_AML_CBFGLIS.model_HH_WNT_TGFB_enrichment_GSEA.csv", row.names = FALSE)
```


### Heatmaps 

```{r}
library(gageData)
library(gage)

data(egSymb)
data(sigmet.idx.hs)
data(kegg.sets.hs)
kegg.all <- kegg.sets.hs[sigmet.idx.hs]
kegg.all.sym <- lapply(kegg.all, eg2sym)

# names(kegg.all.sym)
# kegg.all.sym$`hsa00232 Caffeine metabolism`
```

```{r}
heatmap.samps <- CBFGLIS_samples %>% 
  filter(!grepl("pool", Sample)) %>% 
  #CD34_PB|CBFA2T3|NBM
  filter(grepl("EC_Week1_GFP|EC_Week(1|6|12)_CBFGLIS|Myeloid_Week1_GFP|Myeloid_Week(1|6)_CBFGLIS",
               Comparisons)) %>% 
  #"CD34_PB", "NBM","CBFA2T3-GLIS2",
  mutate(Comparisons=factor(Comparisons, levels=c(
                                                    "EC_Week1_GFP_CordBlood","EC_Week1_CBFGLIS_CordBlood",
                                                    "EC_Week6_CBFGLIS_CordBlood","EC_Week12_CBFGLIS_CordBlood",
                                                    "Myeloid_Week1_GFP_CordBlood", "Myeloid_Week1_CBFGLIS_CordBlood",
                                                    "Myeloid_Week6_CBFGLIS_CordBlood"))) %>%
  arrange(desc(Comparisons)) %>%
  set_rownames(.$Sample)
  
table(heatmap.samps$Comparisons)
# head(heatmap.samps)
# dim(heatmap.samps)
# cc_heatmaps <- colorCodes_aheatmap(df=select(heatmap.samps,all_of(Cols)))
# cc_heatmaps
```

```{r}
paths.to.use <- kegg.all.sym[grep("Hedge|TGF-beta|Wnt", names(kegg.all.sym))]

#Clean up the HH pathway
uniq.HH <- setdiff(paths.to.use$`hsa04340 Hedgehog signaling pathway`, paths.to.use$`hsa04310 Wnt signaling pathway`) %>% 
  setdiff(., paths.to.use$`hsa04350 TGF-beta signaling pathway`) %>% 
  grep("CSNK", ., value=TRUE, invert = TRUE)
paths.to.use$`hsa04340 Hedgehog signaling pathway` <- paths.to.use$`hsa04340 Hedgehog signaling pathway`[paths.to.use$`hsa04340 Hedgehog signaling pathway` %in% uniq.HH]


#clean up the WNT pathwa
# uniq.WNT <- setdiff(paths.to.use$`hsa04310 Wnt signaling pathway`, paths.to.use$`hsa04340 Hedgehog signaling pathway`) %>% 
#   setdiff(., paths.to.use$`hsa04350 TGF-beta signaling pathway`)
# paths.to.use$`hsa04310 Wnt signaling pathway` <- paths.to.use$`hsa04310 Wnt signaling pathway`[paths.to.use$`hsa04310 Wnt signaling pathway` %in% uniq.WNT]
paths.to.use$`hsa04310 Wnt signaling pathway` <- paths.to.use$`hsa04310 Wnt signaling pathway`[grep("^WNT", paths.to.use$`hsa04310 Wnt signaling pathway`)]


#Clean up TGFB path
# uniq.TGFB <- setdiff(paths.to.use$`hsa04350 TGF-beta signaling pathway`, paths.to.use$`hsa04340 Hedgehog signaling pathway`) %>% 
#   setdiff(., paths.to.use$`hsa04310 Wnt signaling pathway`)
# paths.to.use$`hsa04350 TGF-beta signaling pathway` <- paths.to.use$`hsa04350 TGF-beta signaling pathway`[paths.to.use$`hsa04350 TGF-beta signaling pathway` %in% uniq.TGFB]
paths.to.use$`hsa04350 TGF-beta signaling pathway` <- paths.to.use$`hsa04350 TGF-beta signaling pathway`[grep("^BMP", paths.to.use$`hsa04350 TGF-beta signaling pathway`)]


paths.to.use
```

```{r}
cts_models <- in_counts[,heatmap.samps$Sample]
cts_models <- cts_models[rowSums(cts_models) > 10, ]

dds_models <- DESeq2::DESeqDataSetFromMatrix(round(cts_models, digits = 0),
                                      colData = heatmap.samps,
                                      design = ~ 1)


vst_models <- DESeq2::vst(dds_models, blind = TRUE)
dim(vst_models)


in_vst_models <- SummarizedExperiment::assay(vst_models)

dim(in_vst_models)
head(in_vst_models[,1:5])
```

Order:
CBF EC,  CBF MC, GFP EC and GFP MC. 

```{r}
col.order <- in_vst_models %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  filter(gene %in% unlist(paths.to.use)) %>% 
  gather(Sample, VST, -gene) %>% 
  left_join(., select(heatmap.samps,Sample, Comparisons),
            by="Sample") %>% 
  mutate_at(vars(Comparisons), ~factor(., levels=rev(c(
                                                   "EC_Week12_CBFGLIS_CordBlood",
                                                   "EC_Week6_CBFGLIS_CordBlood",
                                                   "EC_Week1_CBFGLIS_CordBlood",
                                                     
                                                   "Myeloid_Week6_CBFGLIS_CordBlood",
                                                   "Myeloid_Week1_CBFGLIS_CordBlood", 
                                                   
                                                   "EC_Week1_GFP_CordBlood",
                                                   "Myeloid_Week1_GFP_CordBlood")))) %>% 
  
  left_join(., rownames_to_column(data.frame(gene=unlist(paths.to.use)), "pathway") %>% 
              mutate(pathway=gsub("[0-9]{1,3}$", "", pathway)), 
            by="gene") %>% 
  arrange(Comparisons, pathway) %>% 
  
  group_by(Comparisons, pathway) %>%
  arrange(desc(VST)) %>%
  ungroup()  %>%
  
  select(Sample,Comparisons, pathway) %>% 
  distinct() %>% 
  arrange(Comparisons)
  
col.order


# col.order %>%
#   filter(path=="hsa04340 Hedgehog signaling pathway")


```


```{r fig.height=5, fig.width=17}
heatmaps <- lapply(names(paths.to.use), function(path){
  
  col_order <- col.order %>%
          filter(pathway==path)

  mat <- in_vst_models[intersect(rownames(in_vst_models),paths.to.use[[path]]), col_order$Sample]

  # mat <- logCPM[intersect(rownames(logCPM), paths.to.use[[path]]), heatmap.samps$Sample]
  # mat <- logCPM_models[intersect(rownames(logCPM_models), paths.to.use[[path]]), heatmap.samps$Sample]
  
  
  len <- 299
  col <- colorRampPalette(c("deepskyblue4", "deepskyblue3", "deepskyblue2", "deepskyblue1","white",
                            "firebrick1", "firebrick2", "firebrick3", "firebrick4"))(n=len)
  col <- colorRampPalette(c("cyan1", "cyan2", "cyan3", "cyan4","black",
                            "magenta4", "magenta3", "magenta2", "magenta1"))(n=len)

  
  p.pathway <- pheatmap::pheatmap(mat=mat,
                       color=col,
                       scale = "row",
                       border_color="black",
                       # breaks = Breaks,
                       cluster_rows=TRUE,
                       cluster_cols = FALSE,
                       clustering_method="ward.D2",
                       clustering_distance_cols="euclidean",
                       annotation_names_col=TRUE,
                       annotation_col  = heatmap.samps[,c("Culture_condition",
                                                         "Transduction_condition",
                                                         "Time_Point_Group")],
                       annotation_colors=list("Culture_condition"=c("EC"="blue1","Myeloid"="red1"),
                                                                    # "NBM"="black", "CD34_PB"="grey40", "CBFA2T3-GLIS2"="mediumorchid"),
                                              "Transduction_condition"=c("GFP"="yellowgreen","CBFA2T3_GLIS"="lightseagreen"),
                                                                         # "NBM"="black", "CD34_PB"="grey40", "CBFA2T3-GLIS2"="mediumorchid"),
                                              "Time_Point_Group"=c("Week1"="steelblue1","Week6"="steelblue3", "Week12"="steelblue4")),
                                                                   # "NBM"="black", "CD34_PB"="grey40", "CBFA2T3-GLIS2"="mediumorchid")),
                       fontsize = 12,
                       fontsize_row=12,
                       fontsize_col=12,
                       filename=paste0("Figures/Heatmaps/Pathway_Genes/TARGET_AML_CBFGLIS.models_",gsub(" ","_",path),"_vst_.pdf"),
                       height = 5,
                       width = 17,
                       show_colnames=FALSE,
                       show_rownames = TRUE)

  
})
```




# GSVA

```{r}
library(GSVA)
```

```{r}
hallmark <- read.gmt("References/h.all.v7.2.symbols.gmt") #h.all.v7.2.symbols.gmt
length(hallmark)
hallmark$HALLMARK_HEDGEHOG_SIGNALING
# table(hallmark$HALLMARK_HEDGEHOG_SIGNALING %in% rownames(logCPM))


hallmark$HALLMARK_HEDGEHOG_SIGNALIN[!hallmark$HALLMARK_HEDGEHOG_SIGNALING %in% rownames(logCPM)]

library(gageData)
library(gage)

data(egSymb)
data(sigmet.idx.hs)
data(kegg.sets.hs)
kegg.sigmet <- kegg.sets.hs[sigmet.idx.hs]
kegg.sigmet.sym <- lapply(kegg.sigmet, eg2sym)
```

```{r}
# logCPM
gsva.res.all <- gsva(expr = logCPM,
                 gset.idx.list = c(hallmark, kegg.sigmet.sym),
                 # annotation=,
                 method="ssgsea",
                 kcdf="Gaussian",
                 parallel.sz=2, 
                 mx.diff=TRUE,
                 abs.ranking=FALSE, 
                 tau=1,
                 min.sz=20,
                 max.sz=400,
                 verbose=TRUE)
```

```{r}
gsva.res.all <- read.csv("GSVA/TARGET_AML_CBFGLIS.models_patients_NBM_ssGSEA_scores.csv", row.names = 1)
# write.csv(gsva.res.all, "GSVA/TARGET_AML_CBFGLIS.models_patients_NBM_ssGSEA_scores.csv")

dim(gsva.res.all) #178 182
head(gsva.res.all[,1:5])
```


```{r}
cc.cultures <- c("CBFA2T3-GLIS2"="mediumorchid2",
                 # "CD34_PB"="grey40",
                 "NBM"="black",
                 # "OtherAML"="grey80",
                 "EC_Week1_GFP_CordBlood"="springgreen4",
                 "EC_Week1_CBFGLIS_CordBlood"="blue2",
                 "EC_Week6_CBFGLIS_CordBlood"="blue2",
                 "EC_Week12_CBFGLIS_CordBlood"="blue2", 
                 "Myeloid_Week1_GFP_CordBlood"="orange2", 
                 "Myeloid_Week1_CBFGLIS_CordBlood"="deeppink", 
                 "Myeloid_Week6_CBFGLIS_CordBlood"="deeppink")
```

```{r}
heatmap.samps <- CBFGLIS_samples %>% 
  filter(!grepl("pool|sort|pos|D63.EC", Sample),
         !c(grepl("replicate", Sample) & Group=="NBM"),
         !grepl("mock",Sample)) %>% 
  filter(!grepl("(Week6|Week12)_GFP|CD34_PB", Comparisons)) %>% 
  mutate_at(vars(Comparisons), ~factor(.,  levels=c(
                                                    # "CD34_PB",
                                                    "EC_Week1_GFP_CordBlood",
                                                    "EC_Week1_CBFGLIS_CordBlood",
                                                    "EC_Week6_CBFGLIS_CordBlood",
                                                    "EC_Week12_CBFGLIS_CordBlood", 
                                                    "Myeloid_Week1_GFP_CordBlood", 
                                                    "Myeloid_Week1_CBFGLIS_CordBlood", 
                                                    "Myeloid_Week6_CBFGLIS_CordBlood",
                                                    "Myeloid_Week12_CBFGLIS_CordBlood",
                                                    "CBFA2T3-GLIS2",
                                                    "NBM"))) %>% 
  filter(!grepl("Myeloid_Week12_CBFGLIS_CordBlood", Comparisons)) %>% 
  droplevels() %>% 
  mutate(Facet=factor(ifelse(grepl("CD34_PB|NBM|CBFA2T3-GLIS2",Comparisons), "Primary Samples", "CBFA2T3-GLIS2 HSPCs"), 
                        levels=c("Primary Samples", "CBFA2T3-GLIS2 HSPCs"))) %>% 
  
  mutate(Label=gsub("TARGET.+((CBFA2T3_GLIS2|GFP).+)$","\\1", Sample)) %>% 
  arrange(desc(Comparisons)) %>%
  set_rownames(.$Sample)


table(heatmap.samps$Comparisons)
# dim(heatmap.samps)
```

```{r}
gsva.res.TF.df <- gsva.res.all[,heatmap.samps$Sample] %>% 
  as.data.frame() %>% 
  rownames_to_column("path") %>% 
  gather(Sample, Score,-path) %>% 
  inner_join(., heatmap.samps, by="Sample") %>%
  filter(grepl("HEDGEHOG|WNT|TGF", path, ignore.case = T)) %>% 
  filter(grepl("hsa", path)) %>% 
  mutate(pathway=path,
         path=gsub("hsa[0-9]+\\s", "", path)) %>% 
  mutate(path=factor(path,levels=c("Hedgehog signaling pathway","TGF-beta signaling pathway","Wnt signaling pathway"))) %>% 
  arrange(path) %>% 
  select(pathway,path, everything())

head(gsva.res.TF.df)
# table(gsva.res.TF.df$NUP98.Rearranged.Groups)
# length(unique(gsva.res.TF.df$path))
# table(heatmap.samps$Comparisons)

# 
# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFA2T3-GLIS2 Associated Paths")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFA2T3-GLIS2 Associated Paths",
#                     dplyr::select(gsva.res.TF.df, Sample,
#                                   pathway_info=pathway,
#                                   path_name=path,
#                                   Enrichemnt_Score=Score,
#                                   AML_Subtype=Comparisons),
#                     keepNA=FALSE,
#                     rowNames=FALSE)


```



```{r fig.height=6, fig.width=25}
# breaks <- plyr::round_any(seq(min(gsva.res.TF.df$Score), max(gsva.res.TF.df$Score), length.out=4), accuracy=0.05)
gsva_breaks_list <- list("Wnt signaling pathway"=c(0.60,0.64,0.68,0.72),
                         "TGF-beta signaling pathway"=c(0.58, 0.62, 0.66, 0.70),
                         "Hedgehog signaling pathway"=c(0.42,0.48,0.54,0.60))

gsva_boxplots <- lapply(as.character(unique(gsva.res.TF.df$path)), function(x){
  
  input <- droplevels(gsva.res.TF.df[gsva.res.TF.df$path==x,])

  # breaks <-  plyr::round_any(seq(min(input$Score), max(input$Score), length.out=5), accuracy=0.02)
  breaks <- gsva_breaks_list[[x]]

  make_gsva_facet_boxplots(gsva_res = input,
                      breaks=breaks)
  
  # return(input)
})


# table(is.na(gsva_boxplots[[1]]$Score))

# gsva_boxplots

# pdf("Figures/Boxplots/TARGET_AML_CBFGLIS.models_CBFGLIS.patients_ssGSEA_boxplots.pdf", height=6, width = 25)
gridExtra::grid.arrange(grobs=gsva_boxplots, nrow=1)
# dev.off()
```

```{r fig.width=15, fig.height=6}
my_comparisons <- list(c("EC_Week1_GFP_CordBlood", "EC_Week1_CBFGLIS_CordBlood"),
                       c("EC_Week1_GFP_CordBlood", "EC_Week6_CBFGLIS_CordBlood"),
                       c("EC_Week1_GFP_CordBlood", "EC_Week12_CBFGLIS_CordBlood"),
                       c("Myeloid_Week1_GFP_CordBlood", "Myeloid_Week1_CBFGLIS_CordBlood"),
                       c("Myeloid_Week1_GFP_CordBlood", "Myeloid_Week6_CBFGLIS_CordBlood"),
                       c("NBM","CBFA2T3-GLIS2"))


ggplot(data=gsva.res.TF.df, aes(x=Comparisons, y=Score, fill=Comparisons)) +
      geom_boxplot(alpha=0.2,outlier.shape = NA,  inherit.aes = TRUE)  +
  stat_compare_means(comparisons = my_comparisons, method="t.test", 
                     method.args = list(alternative="less"), 
                     label="p.signif") + #"p.signif"
  facet_wrap(~path) +
  theme(axis.text.x = element_text(angle=25, vjust=1, hjust=1))
```




### Significant Pathway enrichment

```{r}
#DE analysis with the gsva matrix.
gsva_DE <- function(gsva_matrix,clinData,col,p.value=0.001){
  #clinData must be factor leveled! 
  
  #Example:
  # X <-  sample_info[samps_all, ] %>% 
  #   mutate(USI=Sample) %>%
  #   mutate_at(vars(NUP98.Rearranged.Groups), 
  #             ~factor(., levels=c("OtherAML","NUP98-X"))) %>%
  #   set_rownames(.$Sample)
  # gsva_DE(gsva_matrix = gsva.res.all[,X$Sample],
  #         clinData=X, 
  #         col="NUP98.Rearranged.Groups",
  #         p.value = 0.25)
  
  
  library(edgeR)
  library(limma)
  
  gsva <- gsva_matrix[,clinData$Sample]
  design <- model.matrix(formula(paste("~0 +", col)),
                         data=clinData)
  colnames(design) <- c("Ref","Comparitor")
  
  print("this one")
  
  
  cm <- makeContrasts("Comparitor-Ref", levels = design)
  
  
  fit <- lmFit(gsva, design = design)
  fit2 <- contrasts.fit(fit, contrasts = cm)
  fit2 <- eBayes(fit2)
  
  allGeneSets <- topTable(fit2, p.value = p.value,number=Inf) %>% 
    rownames_to_column("GeneSet") 
    #object 'logFC' not found ℹ Input `..1` is `logFC???? And this error only happens in a lapply()
    # dplyr::arrange(desc(logFC), desc(adj.P.Val))
  
  res <- list("contrast"=cm,"fit"=fit2, "gene_sets"=allGeneSets)
  
  return(allGeneSets)
}

```

#### EC

```{r}
EC.comparisons <-  CBFGLIS_samples %>% #CBFGLIS.models %>% 
  filter(!grepl("sort", Sample), !grepl("Week9", Time_Point_Group)) %>% 
  filter(grepl("EC", Culture_condition)) %>% 
  filter(grepl("GFP|CBFA2T3", Transduction_condition)) %>%
  mutate(EC_Groups_GeneTargets=case_when(
      grepl("EC_Week1_GFP", Comparisons) ~ "GFP", 
      grepl("Week(6|12)_CBFGLIS", Comparisons) ~ "CBFGLIS")) %>% 
  mutate(USI=Sample) %>% 
  set_rownames(.$Sample)

# table(EC.comparisons$Transduction_condition, EC.comparisons$Time_Point_Group)
table(EC.comparisons$Comparisons)
# table(EC.comparisons$EC_Groups_GeneTargets)
# head(EC.comparisons[,1:5])
```

```{r}
ref <- "GFP"
conditions <- c("Week1","Week6", "Week12")

DE.Paths.EC <- lapply(conditions,function(condition){
  df <- filter(EC.comparisons, grepl(ref,Transduction_condition) | 
         (grepl("GLIS",Transduction_condition) & Time_Point_Group == condition)) %>% 
    mutate(Transduction_condition=factor(Transduction_condition,
                                         levels=c(rev(names(table(.$Transduction_condition)))))) %>% 
    set_rownames(.$Sample)
  
  
  levels(df$Transduction_condition)
  table(df$Transduction_condition)
  
  input <- gsva.res.all[grep("hsa", rownames(gsva.res.all)),df$Sample]
  # dim(input)
  
  DE <- gsva_DE(gsva_matrix = input,
                  clinData=df,
                  col="Transduction_condition",
                  p.value = 1)
  
  return(DE)
})

names(DE.Paths.EC) <- conditions


# DE.Paths.EC
```

ns: p > 0.05
*: p <= 0.05
**: p <= 0.01
***: p <= 0.001
****: p <= 0.0001

```{r}
EC.paths.selected <- lapply(names(DE.Paths.EC), function(x)   
  
  dplyr::filter(DE.Paths.EC[[x]], grepl("Hedge|wnt|tgf",GeneSet, ignore.case = T)) %>% 
         dplyr::select(GeneSet ,  matches("adj.P.Val")) %>% 
         mutate(Culture="EC",
                Time_point_Group=x,
                Comparison="C/G-CB vs GFP-CB")
  
  ) %>%
  bind_rows() %>% 
  mutate(adj.P.Val=round(adj.P.Val, digits = 3)) %>% 
  mutate(Signif_level=case_when(
                  adj.P.Val > 0.05 ~ "n.s.",
                  adj.P.Val <= 0.0001 ~ "****",
                  adj.P.Val <= 0.001 ~ "***",
                  adj.P.Val <= 0.01 ~ "**",
                  adj.P.Val < 0.05 ~ "*",
                ))

# EC.paths.selected
# write.csv(EC.paths.selected,"GSVA/TARGET_AML_CBFGLIS.models_vs_GFP_EC_pathways_of_interest.csv", row.names = FALSE)
```


#### MC

```{r}
MC.comparisons <- CBFGLIS_samples %>% 
  filter(!grepl("scbulkleftovers|pool", Sample)) %>% 
  filter(grepl("Myeloid", Culture_condition)) %>% 
  filter(grepl("GFP|CBFA2T3", Transduction_condition)) %>% 
  filter(!c(grepl("GFP", Transduction_condition) & grepl("Week12|Week6", Time_Point_Group))) %>% 
  mutate(USI=Sample) %>% 
  set_rownames(.$Sample)

table(MC.comparisons$Transduction_condition, MC.comparisons$Time_Point_Group)

table(MC.comparisons$Comparisons)
# View(MC.comparisons)
```

```{r}
ref <- "GFP"
conditions <- c("Week1","Week6", "Week12")

DE.Paths.MC <- lapply(conditions,function(condition){
  df <- filter(MC.comparisons, grepl(ref,Transduction_condition) | 
         (grepl("GLIS",Transduction_condition) & Time_Point_Group == condition)) %>% 
    mutate(Transduction_condition=factor(Transduction_condition,
                                         levels=c(rev(names(table(.$Transduction_condition)))))) %>% 
    set_rownames(.$Sample)
  
  
  levels(df$Transduction_condition)
  table(df$Transduction_condition)
  
  input <- gsva.res.all[grep("hsa", rownames(gsva.res.all)),df$Sample]
  # dim(input)
  
  DE <- gsva_DE(gsva_matrix = input,
                  clinData=df,
                  col="Transduction_condition",
                  p.value = 1)
  
  return(DE)
})

names(DE.Paths.MC) <- conditions
# DE.Paths.EC
```

```{r}
MC.paths.selected <- lapply(names(DE.Paths.MC), function(x)   
  
  filter(DE.Paths.MC[[x]], grepl("Hedge|wnt|tgf",GeneSet, ignore.case = T)) %>% 
         dplyr::select(GeneSet ,  matches("adj.P.Val")) %>% 
         mutate(Culture="MC",
                Time_point_Group=x,
                Comparison="C/G-CB vs C/G-GFP")
  
  ) %>%
  bind_rows() %>% 
  mutate(adj.P.Val=round(adj.P.Val, digits = 3)) %>% 
  mutate(Signif_level=case_when(
                  adj.P.Val > 0.05 ~ "n.s.",
                  adj.P.Val <= 0.0001 ~ "****",
                  adj.P.Val <= 0.001 ~ "***",
                  adj.P.Val <= 0.01 ~ "**",
                  adj.P.Val < 0.05 ~ "*",
                ))

MC.paths.selected
# write.csv(MC.paths.selected,"GSVA/TARGET_AML_CBFGLIS.models_vs_GFP_MC_pathways_of_interest.csv", row.names = FALSE)
```


```{r}
gsva_sig_res <- bind_rows(EC.paths.selected, MC.paths.selected)


# openxlsx::addWorksheet(supp_table,
#                        sheetName = "HH, WNT, BMP ssGSEA Enrichment")
# openxlsx::writeData(supp_table,
#                     sheet = "HH, WNT, BMP ssGSEA Enrichment",
#                     gsva_sig_res,
#                     keepNA=FALSE,
#                     rowNames=FALSE)
```



#Check for Contaminants 

```{r}
#EC marker genes
# markers <- c( "CD144", "CD31")
markers <- c("CDH5","PECAM1","RAI2","STXBP6","SLC30A3")

filter(geneIDmap, gene_name %in% markers)
```

```{r}
input_samples <- CBFGLIS_samples %>% 
  filter(!grepl("pool", Sample)) %>% 
  #CD34_PB|CBFA2T3|NBM
  filter(grepl("CBFA2T3|NBM", Comparisons) | 
           grepl("EC_Week1_GFP|EC_Week(1|6|12)_CBFGLIS|Myeloid_Week1_GFP|Myeloid_Week(1|6)_CBFGLIS", Comparisons)) %>% 
  
  #"CD34_PB", "NBM","CBFA2T3-GLIS2",
  mutate(Comparisons=factor(Comparisons, levels=c( "NBM","CBFA2T3-GLIS2",
                                                    "EC_Week1_GFP_CordBlood",
                                                   "EC_Week1_CBFGLIS_CordBlood",
                                                   "EC_Week6_CBFGLIS_CordBlood",
                                                   "EC_Week12_CBFGLIS_CordBlood",
                                                   
                                                    "Myeloid_Week1_GFP_CordBlood",
                                                   "Myeloid_Week1_CBFGLIS_CordBlood",
                                                    "Myeloid_Week6_CBFGLIS_CordBlood"))) %>%
  arrange(desc(Comparisons)) %>%
  set_rownames(.$Sample)
  
table(input_samples$Comparisons)
```

```{r}
TPM.markers <- TPM[rownames(TPM) %in% markers, input_samples$Sample] %>% 
  rownames_to_column("gene") %>% 
  gather(Sample, TPM, -gene) %>% 
  mutate(log2_TPM=log2(TPM+1)) %>% 
  left_join(., input_samples, by="Sample")

head(TPM.markers)
```

```{r fig.width=20, fig.height=5}
# pdf("TARGET_AML_SLC30A3_expression_boxplots.pdf", height = 5, width = 20)
ggplot(filter(TPM.markers, grepl("SLC30A3", gene)),
       aes(x=Comparisons, y=log2_TPM, fill=Comparisons)) +
  geom_boxplot() +
  facet_wrap(~gene, scales="free") +
  theme_classic() +
  theme(axis.text.x = element_text(angle=25, vjust=1, hjust=1))

# dev.off()
```

```{r}
input_samples %>% 
  group_by(Comparisons) %>% 
  count() %>% 
  write.csv(., "TARGET_AML_CBFGLIS.models_Samples_Table.csv", row.names = FALSE)
```

# Target Genes 

```{r}
#Read in 
all_targets <- read.delim("Theraputic_Targets/ALL_GENES.output.candidate_genes.gt50percent_gt2stdev_for1subtype.v3.tsv",
                          sep="\t", header = FALSE) %>% 
  rename_all(~c("gene_name","Enriched_Groups")) %>%
  left_join(., geneIDs, by="gene_name") %>%
  left_join(., select(geneIDmap,-gene_name),
            by="gene_id") %>% 
  mutate(CBFGLIS.Enriched=case_when(
    grepl("cbfa2t3-glis2", Enriched_Groups) ~ "Yes",
    TRUE ~ "No"))


head(all_targets)
dim(all_targets)

# table(all_targets$CBFGLIS.Enriched) #299 enriched in CBFGLIS


#Read in the DEGs results
DEGs.EC.week7 <- read.csv("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week7_EC.culture_DEGs_5.21.21.csv", comment.char = "#")


DEGs.EC.week12 <- read.csv("DEGs/TARGET_AML_CBFGLIS.model_vs_GFP_Week12_EC.culture_DEGs_5.21.21.csv",comment.char = "#")

```

```{r}
samples_to_use <- CBFGLIS_samples %>% 
  filter(!grepl("pool", Sample)) %>%
  filter(!grepl("_replicate", Sample)) %>% 
  #CD34_PB|CBFA2T3|NBM
  filter(grepl("CBFA2T3|NBM|CD34_PB", Comparisons) | 
           grepl("EC_Week1_GFP|EC_Week(6|12)_CBFGLIS", Comparisons)) %>% 
  
  #"CD34_PB", "NBM","CBFA2T3-GLIS2",
  mutate(Comparisons=factor(Comparisons, levels=c( "NBM",
                                                   "CD34_PB",
                                                   "CBFA2T3-GLIS2",
                                                    "EC_Week1_GFP_CordBlood",
                                                   # "EC_Week1_CBFGLIS_CordBlood",
                                                   "EC_Week6_CBFGLIS_CordBlood",
                                                   "EC_Week12_CBFGLIS_CordBlood"))) %>% 
                                                   
                                                   #  "Myeloid_Week1_GFP_CordBlood", 
                                                   # "Myeloid_Week1_CBFGLIS_CordBlood",
                                                   #  "Myeloid_Week6_CBFGLIS_CordBlood"))) %>%
  arrange(desc(Comparisons)) %>%
  mutate(Group=case_when(
    grepl("NBM|CD34_PB", Comparisons) ~ "Normals",
    TRUE ~ as.character(Comparisons))) %>% 
  set_rownames(.$Sample)
  
table(samples_to_use$Comparisons)
table(samples_to_use$Group)
dim(samples_to_use)
```


```{r}
# volcano_plots_list <- dir("DEGs/for_volcano_plots/", full.names = TRUE)
# 
# volcano_plots_dfs <- lapply(volcano_plots_list, read.csv, comment.char="#")
# names(volcano_plots_dfs) <-  dir("DEGs/for_volcano_plots/")
# lapply(volcano_plots_dfs, head)
```

## Filter for Absent Expression

```{r}
#Calculate summary statistics for target genes
TPM.ratio.res <- TPM[rownames(TPM) %in% all_targets$gene_name,samples_to_use$Sample] %>% 
  rownames_to_column("gene") %>% 
  gather(Sample, TPM, -gene) %>% 
  
  left_join(., samples_to_use, by="Sample") %>% 
  group_by(gene, Group) %>% 
  summarize(min=min(TPM),
            median=median(TPM),
            mean=mean(TPM),
            max=max(TPM),
            percent_expressors=round((sum(TPM >= 1.0)/n())*100, digits = 2)) %>% 
  mutate(Range=max-min) %>% 
  ungroup() %>% 
  rename_at(vars(min:Range), ~paste0(.,"_TPM")) %>% 
  
  gather(Stat,Value, -gene, -Group) %>%
  pivot_wider(id_cols = gene,
              names_from=c(Group,Stat),
              values_from=Value) %>%

  select(colnames(.)[order(colnames(.))]) %>%
  select(gene, matches("^Normals"), matches("^NBM"),everything())


head(TPM.ratio.res)
dim(TPM.ratio.res) #993  31
```

```{r}
TPM.ratio.filt <- TPM.ratio.res %>% 
  filter(Normals_max_TPM < 1.0)

dim(TPM.ratio.filt) #607
# table(samples_to_use$Comparisons)
```


## Filter for CBFGLIS Over-expression

```{r}
all_inputs <- list(all_targets=TPM.ratio.filt$gene,
                   DEGs.EC.week7=DEGs.EC.week7$gene[DEGs.EC.week7$logFC > 0],
                   DEGs.EC.week12=DEGs.EC.week12$gene[DEGs.EC.week12$logFC > 0],
                   CBFGLISvsNBM=CBFGLISvsNBM$gene_name[CBFGLISvsNBM$logFC > 0])

lapply(all_inputs, head)
sapply(all_inputs, length)
```

```{r}
res_4_inputs <- multi_intersect(l=all_inputs)
res_4_inputs <- res_4_inputs[order(res_4_inputs)]


# res_4_inputs
length(res_4_inputs) #50 genes


# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CBFA2T3-GLIS2 DEGs vs NBM")
# openxlsx::writeData(supp_table,
#                     sheet = "CBFA2T3-GLIS2 DEGs vs NBM",
#                     CBFGLISvsNBM,
#                     keepNA=FALSE,
#                     rowNames=FALSE)

```

```{r}
#Subset the DEGs results
targets.df <- lapply(names(all_inputs)[-1], function(x){
  
  dataset <- gsub("DEGs.", "", gsub("7", "6", x))
  input <- get(x)
  res <- input %>%
    dplyr::select(matches("^gene$|^gene_name$"), logFC, adj.P.Val) %>%
    rename_at(vars(logFC:adj.P.Val), ~paste(., dataset, sep="_"))


  if(x=="CBFGLISvsNBM"){
    res$gene <- res$gene_name
    res <- res[,-grep("gene_name", colnames(res))]
  }


  res <- res %>%
    filter(gene %in% res_4_inputs)

  return(res)
  
})

#join the dataframes
targets.df <- Reduce(function(x, y, ...) inner_join(x, y, by="gene"), targets.df) %>%
  left_join(., all_targets, by=c("gene"="gene_name")) %>%
  dplyr::select(gene, Cell_Surface_Protein, Cell_Adhesion_Gene,  Transmembrane_Helix, gene_type, everything()) %>%
  arrange(desc(Cell_Surface_Protein), desc(logFC_EC.week12))

#Calculate summary statistics for target genes
TPM.targets <- TPM[rownames(TPM) %in% targets.df$gene,samples_to_use$Sample] %>% 
  rownames_to_column("gene") %>% 
  gather(Sample, TPM, -gene) %>% 
  
  left_join(., samples_to_use, by="Sample") %>% 
  group_by(gene, Comparisons) %>% 
  summarize(min=min(TPM),
            median=median(TPM),
            mean=mean(TPM),
            Q3=quantile(TPM, probs=0.75),
            max=max(TPM),
            percent_expressors=round((sum(TPM >= 1.0)/n())*100, digits = 2)) %>% 
  mutate(Range=max-min) %>% 
  ungroup() %>% 
  rename_at(vars(min:Range), ~paste0(.,"_TPM")) %>% 
  
  gather(Stat,Value, -gene, -Comparisons) %>%
  pivot_wider(id_cols = gene,
              names_from=c(Comparisons,Stat),
              values_from=Value) %>%
  
  #Filter to ensure that genes are not also expressed in the GFP controls
  filter(EC_Week1_GFP_CordBlood_max_TPM < 1.0) %>%

  #Reoder the columns 
  select(colnames(.)[order(colnames(.))]) %>%
  select(gene, matches("^NBM"),everything()) %>% 
  rename_at(vars(matches("CBFA2T3-GLIS2")), ~gsub("-", "\\.", .))


# options(scipen = 999)
# dim(TPM.targets)
# head(TPM.targets)

#Merge in the summary stats
targets.df <- targets.df %>% 
  inner_join(., TPM.targets, by="gene") %>% 
  select(gene:adj.P.Val_CBFGLISvsNBM, matches("^NBM|CBFA2T3|EC"), 
         everything())


# head(targets.df)
dim(targets.df) # 42 67 
table(targets.df$Cell_Surface_Protein)
# write.csv(targets.df, "Theraputic_Targets/TARGET_AML_CBFGLIS.models_AMLrestrictedGene_Targets_SummaryStats_07.06.21.csv", row.names = FALSE)
```


## Filter for expression in CBFGLIS 

```{r}
quantile(targets.df$EC_Week6_CBFGLIS_CordBlood_Range_TPM)
quantile(targets.df$EC_Week12_CBFGLIS_CordBlood_Range_TPM)



# targets.df %>%
#   filter(gene=="FOLR1") %>% 
#   select(gene, matches("Q3"))

```

```{r}
targets.df.filtered <- targets.df  %>% 
  filter(Cell_Surface_Protein=="Yes") %>% 
  
  #low in NBM, but large amount of expressors in the CBFGLIS CB+
  filter(EC_Week6_CBFGLIS_CordBlood_percent_expressors_TPM >= 75.0, 
         EC_Week12_CBFGLIS_CordBlood_percent_expressors_TPM >= 75.0) %>%  
  
  #dynamic range filter
  filter(EC_Week6_CBFGLIS_CordBlood_max_TPM >= 5.0, 
         EC_Week12_CBFGLIS_CordBlood_max_TPM >= 5.0) %>% 
  
  
  arrange(desc(round(logFC_EC.week12))) %>% 
  mutate(Rank=1:n()) %>%
  select(gene, Rank, everything())


# targets.df.filtered
dim(targets.df.filtered)
targets.df.filtered


# quantile(targets.df.filtered$CBFA2T3.GLIS2_percent_expressors_TPM) # 4 targets have >
# quantile(targets.df.filtered$EC_Week6_CBFGLIS_CordBlood_percent_expressors)
# quantile(targets.df.filtered$EC_Week1_GFP_CordBlood_max)


# write.csv(targets.df.filtered, "Theraputic_Targets/TARGET_AML_CBFGLIS.models_AMLrestrictedGene_Targets_filtered_on_CBmodelsOnly_10.18.21.csv", row.names = FALSE)
```

```{r}
# openxlsx::addWorksheet(supp_table,
#                        sheetName = "CAR-T Targets")
# openxlsx::writeData(supp_table,
#                     sheet = "CAR-T Targets",
#                     dplyr::select(targets.df.filtered,-c(Cellular.Compartment_Membrane:CBFGLIS.Enriched),
#                                   -Enriched_Groups, -Ensembl_ProteinID_with_MembraneLocalization),
#                     keepNA=FALSE,
#                     rowNames=FALSE)
## write out the final results
# openxlsx::saveWorkbook(supp_table,
#                        "CBFA2T3.GLIS2_HSPCs_Supplemental_Tables_Raw_Figure_Data.xlsx",
#                        overwrite = T)
```


## Gene Targets Figures 

## Violin plots 

```{r}
samps_for_violin <- CBFGLIS_samples %>% 
  filter(!grepl("pool|sort|pos|D63.EC", Sample),
         !c(grepl("replicate", Sample) & Group=="NBM"),
         !grepl("mock",Sample)) %>% 
  # filter(!grepl("(Week6|Week12)_GFP|CD34_PB", Comparisons)) %>% 
  filter(grepl("CBFA2T3-GLIS2|NBM|CD34_PB", Comparisons)) %>% 
  mutate_at(vars(Comparisons), ~factor(.,  levels=c(
                                                   
                                                    # "EC_Week1_GFP_CordBlood",
                                                    # "EC_Week1_CBFGLIS_CordBlood",
                                                    # "EC_Week6_CBFGLIS_CordBlood",
                                                    # "EC_Week12_CBFGLIS_CordBlood", 
                                                    # "Myeloid_Week1_GFP_CordBlood", 
                                                    # "Myeloid_Week1_CBFGLIS_CordBlood", 
                                                    # "Myeloid_Week6_CBFGLIS_CordBlood",
                                                    # "Myeloid_Week12_CBFGLIS_CordBlood",
                                                    "CBFA2T3-GLIS2",
                                                     "CD34_PB",
                                                    "NBM"))) %>% 
  filter(!grepl("Myeloid_Week12_CBFGLIS_CordBlood", Comparisons)) %>% 
  droplevels() %>% 
  mutate(Facet=factor(ifelse(grepl("CD34_PB|NBM|CBFA2T3-GLIS2",Comparisons), "Primary Samples", "CBFA2T3-GLIS2 HSPCs"), 
                        levels=c("Primary Samples","CBFA2T3-GLIS2 HSPCs")))
  

# table(samps_for_violin$Comparisons, useNA='ifany')
# table(samps_for_violin$Facet, useNA='ifany')
# table(samps_for_violin$Culture_condition, useNA='ifany')
# unique(samps_for_violin$Comparisons[order(samps_for_violin$Comparisons)])



targets.expn <- CPM[targets.df.filtered$gene,samps_for_violin$Sample] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene_name") %>% 
  gather(Sample, CPM, -gene_name) %>% 
  mutate(log2CPM=log2(CPM+1)) %>% 
  left_join(.,samps_for_violin, by="Sample")

# targets.expn
levels(targets.expn$Comparisons)
```

```{r fig.width=10, fig.height=5}

# pdf("TARGET_AML_Immunotheraputic_Targets.pdf", height = 5, width = 10)
ggplot(targets.expn, aes(x=gene_name, y=log2CPM)) +
  geom_point(aes(color=Comparisons), position=position_jitterdodge(jitter.width=0.75)) +
  # geom_boxplot(aes(fill=Comparisons, alpha=0.3)) +
  geom_violin(aes(fill=Comparisons), draw_quantiles = 0.5,
              scale="width", alpha=0.5) +
  scale_color_manual(values=c("CBFA2T3-GLIS2"="dodgerblue4", "NBM"="transparent", "CD34_PB"="transparent"),
                     labels=c("CBFA2T3-GLIS2"="CBFA2T3-GLIS2 AML", "NBM"="Normal Marrows", "CD34_PB"="CD34+ PB")) +
  scale_fill_manual(values=c("CBFA2T3-GLIS2"="dodgerblue4", "NBM"="black", "CD34_PB"="grey80"),
                    labels=c("CBFA2T3-GLIS2"="CBFA2T3-GLIS2 AML", "NBM"="Normal Marrows", "CD34_PB"="CD34+ PB")) +
  labs(x="Immunotheraputic Target Genes", y="Normalized Expression\n(log2 CPM)") +
  theme_classic() +
  theme(axis.text.x = element_text(size=18, color="black"),
        axis.text.y = element_text(size=18, color="black"), 
        axis.title = element_text(size=20), 
        legend.position = c(0.15, 0.9),
        legend.text = element_text(size=14),
        legend.title = element_blank())

# dev.off()
```



## Volcano plots

```{r fig.height=7, fig.width=10}
library(ggrepel)
require(circlize)
library(RColorBrewer)
```

```{r}
degs.cbfglis <- read.csv("DEGs/for_volcano_plots/TARGET_AML_CBFGLIS_vs_NBM_DEGs_forVolcanoPlots.csv", comment.char = "#") %>% 
  mutate(label=ifelse(gene_name %in% targets.df.filtered$gene,gene_name,""),
         Neg.log10_FDR=-log10(FDR))

head(degs.cbfglis)
range(degs.cbfglis$log2FC)
```



```{r fig.height=8, fig.width=8, warning=FALSE}
# colorPal <- colorRampPalette(c("dodgerblue3", "dodgerblue2", "dodgerblue1","white","red1", "red2", "red3", "red4"))(n=299)
# colorPal <- viridis::turbo(299)
# colorPal <- colorRampPalette(rev(brewer.pal(n=10,"RdBu")))(n=20)
#"#D1E5F0",
colorPal <- colorRampPalette(c("#053061", "#2166AC", "#4393C3", "#92C5DE",  "white","#FDDBC7", "#F4A582", "#D6604D", "#B2182B", "#67001F"))(n=20)


# pdf("CBFGLIS_vs_OtherAML_DEGs_VolcanoPlot.pdf", height = 8, width =8)
volcano <- ggplot(degs.cbfglis, aes(x=log2FC, y=Neg.log10_FDR, label = label)) + 
  geom_point(aes(colour= log2FC), size=5, alpha=1) +

  scale_color_gradientn(colours = colorPal,
                        values=scales::rescale(c(-1,-0.075,1)),
                        name="Log2 FC") +
  # scale_color_gradient2(low="#67A9CF", mid="#F7F7F7", high="#EF8A62") +
  # scale_color_gradient2() +
  scale_x_continuous(limits = c(-9,11), breaks = c(-10,6,-2,0,2,6,10)) +
  theme(plot.title = element_text(hjust = 0.5, size = 18),
                       panel.background = element_rect(fill="white"),
                       panel.grid.major = element_blank(),
                       panel.grid.minor = element_blank(),
                       panel.border = element_rect(color = "black", fill=NA),
                       axis.text = element_text(color = "black"),
                       axis.text.x = element_text(angle = 0,hjust=0.5,vjust = 0.5, size = 18),
                       axis.text.y = element_text(size = 18),
                       axis.title = element_text(size = 18),
                       plot.margin = margin(8,2,20,2),
        legend.position = c(0.185,0.94),
        legend.direction = "horizontal",
        legend.text = element_text(size=14),
        legend.title = element_text(size=14, hjust = -0.5),
        legend.key.size =  unit(0.75, "cm")) +
  labs(x="Log2 Fold-Change (FC)", y="-log10(FDR)", title="DE Genes in CBFA2T3-GLIS2\n Diagnostic AML") +
    geom_text_repel(size=6,min.segment.length=unit(0.0001,"mm"),
                    box.padding = unit(0.5, "lines"),
                    point.padding = unit(0.3, "lines"),
                    nudge_x=1, fontface="bold")
 

volcano

# dev.off()
```


## Correlate FCs 
```{r}
targets.df.filtered <- targets.df.filtered %>% 
  mutate(Percent_Expressors=round(`CBFA2T3-GLIS2_percent_expressors_TPM`)) %>% 
  mutate(Percent_Expressors_binned=case_when(
    Percent_Expressors >= 70 & Percent_Expressors < 80 ~ "70-79%",
    Percent_Expressors >= 80 & Percent_Expressors < 90 ~ "80-89%",
    Percent_Expressors >= 90  ~ "90-100%"))


dim(targets.df.filtered)
```

```{r}
targets.df.filtered$NBM_median_TPM
targets.df.filtered$EC_Week12_CBFGLIS_CordBlood_mean_TPM
targets.df.filtered$EC_Week12_CBFGLIS_CordBlood_percent_expressors_TPM

# -log10(targets.df.filtered$adj.P.Val_CBFGLISvsNBM)
targets.df.filtered$`CBFA2T3-GLIS2_mean_TPM`
targets.df.filtered$`CBFA2T3-GLIS2_median_TPM`
```

```{r}
library(RColorBrewer)
gene_colors <- brewer.pal(9,"Set1") 
gene_colors <- gene_colors[-c(5:6)]
gene_colors <- c(gene_colors, "khaki3","cadetblue3") %>%  #"burlywood3"
  set_names(targets.df.filtered$gene)

names(gene_colors)[1] <- "FOLR1"
names(gene_colors)[2] <- "HPSE2"


barplot(rep(1,9), col=gene_colors,names.arg = names(gene_colors), las=2)
```

```{r}
table(targets.df.filtered$Percent_Expressors)
pt_sizes <- c("70-79%"=5, "80-89%"=7, "90-100%"=9)
```

```{r fig.height=7, fig.width=10}
ranked_plot <- ggplot(targets.df.filtered, 
                      aes(x=logFC_EC.week12, 
                          y=logFC_CBFGLISvsNBM,
                          # size=`CBFA2T3-GLIS2_median_TPM`,
                          # shape=Percent_Expressors_binned,
                          color=gene,
                          )) +
  # geom_smooth(data=targets.df.filtered, 
  #             aes(x=logFC_EC.week12, y=logFC_CBFGLISvsNBM),
  #             alpha=0.1,
  #             formula = 'y ~ x', method = lm, 
  #             inherit.aes = FALSE) +
  geom_point(alpha=1.0, size=7) +
  geom_text_repel(aes(label=gene),
                  point.padding = unit(3,"mm"),
                  min.segment.length = unit(5, "mm")) +
  guides(color=FALSE) +
         # size=guide_legend(title="")) + #Percent Expressing Gene\nin CBFA2T3-GLIS2

  # scale_size_continuous(breaks=c(70,80,90,100), range=c(3,10)) +
  # scale_size_continuous(breaks=c(4,6,8,10), range=c(3,7)) +
  # scale_size_manual(values=pt_sizes)) +

  scale_color_manual(values=gene_colors) +
  scale_x_continuous(breaks=c(2,4,6,8,10), limits=c(1,10)) +
  scale_y_continuous(breaks=c(2,4,6,8,10), limits=c(1,10)) +
  
  labs(x="log2 FC  (CBFA2T3-GLIS2 EC Week 12)", y="Log2 FC (CBFA2T3-GLIS2 Patients)",
       title="Theraputic Targets in CBFA2T3-GLIS2 AML") +
  theme_classic() +
  theme(axis.text = element_text(size=12, color="black"),
        axis.title = element_text(size=16, color="black"))

ranked_plot
```


```{r}
# ggsave(plot=ranked_plot, filename = "Figures/TARGET_AML_Gene_Targets_Top10_ScatterPlot_5.21.21.pdf", height = 5, width = 7)
```

```{r}
# filter(targets.df, gene %in% c("NCAM1", "GABRE"))
```

```{r}
nbm.NCAM1 <- TPM["NCAM1",grep("RO[0-9]|BM[0-9]", colnames(TPM))]
nbm.NCAM1[order(nbm.NCAM1, decreasing = T)]
```

# FOLR1 MFI

```{r}
MFI <- read.table("2021_ASH/MFI_Fold_Change_FOLR1.txt",header = TRUE) %>% 
  mutate(Sample=paste0("Sample",1:nrow(.))) %>% 
  gather(Cell_Type, MFI_FC, AML:Myeloid)

head(MFI)
```

```{r}
range(MFI$MFI_FC)
```

```{r fig.height=5, fig.width=4.5}
# pdf("TARGET_AML_FoldChange_MFI_FOLR1_boxplot.pdf", height = 5, width = 4.5)
ggplot(MFI, aes(x=Cell_Type, y=MFI_FC, fill=Cell_Type, color=Cell_Type)) +
  geom_boxplot(alpha=0.3) +
  geom_point(position =   ggbeeswarm::position_quasirandom()) +
  scale_fill_manual(values=c("AML"="dodgerblue4", "Lymph"="black", "Mono"="grey50", "Myeloid"="grey80"),
                    labels=c("AML"="CBFA2T3-GLIS2", "Lymph"="Lymphocytes","Mono"="Monocytes", "Myeloid"="Myeloid")) +
  scale_color_manual(values=c("AML"="dodgerblue4", "Lymph"="black", "Mono"="grey50", "Myeloid"="grey80"),
                    labels=c("AML"="CBFA2T3-GLIS2", "Lymph"="Lymphocytes","Mono"="Monocytes", "Myeloid"="Myeloid")) +
  scale_x_discrete(labels=c("AML"="CBFA2T3-GLIS2", "Lymph"="Lymphocytes","Mono"="Monocytes", "Myeloid"="Myeloid")) +
  labs(y="Fold-Change MFI [AML/Normal]", x="Cell Type") +
  theme_classic() +
    theme(axis.text.x = element_text(size=10, color="black", angle=10, vjust=1, hjust=1),
        axis.text.y = element_text(size=16, color="black"), 
        axis.title = element_text(size=16), 
        legend.position = c(0.75, 0.9),
        legend.text = element_text(size=14),
        legend.title = element_blank())

# dev.off()
```

# FOLR1 vs FOLR2 Expression

```{r}
samps_folr_expn <- sample_info %>% 
  filter(grepl("CBFA2T3-GLIS2|NBM", AML_Subtype))  %>% 
  mutate_at(vars(Time_point), ~case_when(
    Group=="FlowSorted" & grepl("Unsorted.0[39]A",Sample) ~ "diagnostic",
    TRUE ~ .)) %>% 
  mutate_at(vars(Sample), ~case_when(
    grepl("PATGIG|PATISD", .) ~ gsub("_replicate", "", .),
    TRUE ~ .)) %>% 
  filter(!grepl("replicate", Sample) , !grepl("\\.Sorted\\.", Sample), 
         Time_point != "remission") %>% 
  set_rownames(.$Sample)


table(samps_folr_expn$Group)
table(samps_folr_expn$AML_Subtype, samps_folr_expn$Time_point)
# all(samps_folr_expn$Sample %in% colnames(cts)) #TRUE
dim(samps_folr_expn)
```

```{r}
genes <- grep("^ENSG", geneIDs$gene_id)

cts_folr <- cts[genes, samps_folr_expn$Sample]
AML <- !grepl("BM[0-9]|R[O0][0-9]", colnames(cts_folr))
keep <- rowSums(cpm(cts_folr[,AML]) >= 1) >= 0.05*ncol(cts_folr[,AML])

#Filter to remove low/zero count genes
cts_folr <- cts_folr[keep, ]
dge_folr <- DGEList(counts=cts_folr)
dge_folr <- calcNormFactors(dge_folr,method = "TMMwsp")

logCPM_folr <- edgeR::cpm(dge_folr,log=TRUE,normalized.lib.sizes=TRUE, prior.count=1)
head(logCPM_folr[,1:5])
dim(logCPM_folr)
```

reviewer comments:

* Given the high homology between both genes, it would be important to confirm both on the genetic level and protein level, that FOLR1 and not FOLR2, previously identified as target in AML are indeed detected.

* They give an example of 4 patient samples looking for FOLR1 expression, is there a data base where we can know what % of patients AMKL samples have FOLR1 expression?

* In discussion they mention that the treated mice eventually relapsed, did you look at FOLR1 expression after relapse?

```{r}
subset_folr1 <- logCPM_folr[grep("FOLR[12]", rownames(logCPM_folr), value=T),samps_folr_expn$Sample]
dim(subset_folr1)
```

```{r fig.height=3, fig.width=15}
quickPheatmap(expn=subset_folr1, geneList = c("FOLR1","FOLR2"), 
              clinData = samps_folr_expn, 
              cols=c("AML_Subtype", "Time_point"), 
              annots_col_colors=NULL)
```

```{r}
# colors <- c("CBFA2T3-GLIS2"="mediumorchid2",  "NBM"="grey20")

subset_folr1_long <- subset_folr1 %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  pivot_longer(cols=matches("^TARGET"), 
               names_to="Sample", values_to="log2_CPM") %>% 
  group_by(gene) %>% 
  mutate(z_score=scale(log2_CPM)) %>% 
  ungroup() %>% 
  left_join(.,samps_folr_expn,by="Sample") %>% 
  mutate(labels=paste(Time_point, gene)) %>% 
  mutate(Time_point=factor(Time_point, levels=c("diagnostic","relapse","NBM"))) %>% 
  mutate(labels=factor(labels, levels=c("diagnostic FOLR1", 
                   "diagnostic FOLR2",
                   "relapse FOLR1", 
                   "relapse FOLR2",
                   "NBM FOLR1", 
                   "NBM FOLR2")))


# order_samples <- hclust(dist(t(subset_folr1)), method = "ward.D2")

order_samples <- subset_folr1_long %>%
  pivot_wider(id_cols = c(Sample,Time_point),
              names_from=gene, values_from=log2_CPM) %>% 
  mutate_at(vars(FOLR1, FOLR2), ~round(., digits=0)) %>% 
  group_by(Time_point) %>% 
  arrange(Time_point, desc(FOLR1), desc(FOLR2), .by_groups=TRUE) %>% 
  ungroup()

# View(order_samples)

# levels <- order_samples$labels[order_samples$order]
levels <- unique(order_samples$Sample)

subset_folr1_long <- subset_folr1_long %>% 
  mutate(gene=as.factor(gene)) %>%
  mutate(Sample=factor(Sample, 
                       levels=levels))

# subset_folr1_long
```

```{r}
TPM[grep("FOLR1", rownames(TPM)), samps_folr_expn$Sample] %>% 
  as.data.frame() %>% 
  rownames_to_column("gene") %>% 
  pivot_longer(cols=matches("^TARGET"), 
               names_to="Sample", 
               values_to="TPM") %>%
  left_join(.,samps_folr_expn, by="Sample") %>% 
  group_by(Time_point) %>% 
  summarize(Total_N_Samples=n(),
            N_FOLR1_expressors=sum(TPM >= 1)) %>% 
  ungroup() %>% 
  mutate(Percent_FOLR1_expressors=round(N_FOLR1_expressors/Total_N_Samples *100,digits=2))
  # write.csv(.,"TARGET_AML_CBFA2T3-GLIS2_expressors_GT.1TPM_table.csv", row.names = FALSE)
```

```{r fig.height=5, fig.width=15}
annots_df <- samps_folr_expn %>% 
  select(Sample, AML_Subtype, Time_point) %>% 
  pivot_longer(cols=c(AML_Subtype, Time_point), names_to="column", values_to="value") %>% 
  mutate(Sample=factor(Sample, 
                       levels=levels))

# annots_df

colors <- c("CBFA2T3-GLIS2"="mediumorchid2",  "NBM"="grey20", "diagnostic"="burlywood", "relapse"="burlywood4")

annots_plot <- ggplot(annots_df, aes(x=Sample, y=column, fill=value)) +
  geom_col() +
  coord_fixed() +
  scale_fill_manual(values=colors) +
  theme_void() +
  theme(legend.title = element_blank())


pal <- colorRampPalette(c("deepskyblue4", "deepskyblue3", "deepskyblue2", "deepskyblue1",
                          "white","red1", "red2", "red3", "red4"))(n=299)


tile_plot <- ggplot(subset_folr1_long, aes(x=Sample, y=gene, fill=z_score)) + 
  geom_tile(color="black") +
  coord_fixed(ratio=1.5) +
  labs(x="", y="") +
  scale_y_discrete(expand=c(0,0)) +
  # scale_fill_gradient2(  low = "blue", mid = "white", high = "red") +
  scale_fill_gradientn(colors=pal) +
  theme(axis.text.x = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text.y = element_text(size=12))

# pdf("Figures/TARGET_AML_CBFGLIS_diagnostic_relapse_NBM_FOLR1_FOLR2_heatmap.pdf", height = 5, width = 15)
annots_plot / tile_plot + patchwork::plot_layout(guides = "collect")
# dev.off()
```


```{r}
unique(subset_folr1_long$labels)
```

```{r}
min(subset_folr1_long$log2_CPM)
```


```{r fig.height=5, fig.width=10}
# subset_folr1_long
labels_colors <- c("diagnostic FOLR1"="mediumorchid2", 
                   "diagnostic FOLR2"="paleturquoise3",
                   "relapse FOLR1"="mediumorchid4", 
                   "relapse FOLR2"="paleturquoise4",
                   "NBM FOLR1"="black", 
                   "NBM FOLR2"="grey50")

comparisons <- list(c("diagnostic FOLR1", "diagnostic FOLR2"),
                   c("relapse FOLR1","relapse FOLR2"),
                   c("NBM FOLR1", "NBM FOLR2"))

folr_violins <- ggplot(subset_folr1_long, aes(x=labels, y=log2_CPM, fill=labels)) +
  geom_point(aes(color=labels),  size=2.0, alpha=0.7,
             position = position_jitterdodge(jitter.width = 1.25, dodge.width = 1)) +
  geom_violin(draw_quantiles = 0.5, alpha=0.25,scale = "width") +
  labs(x="") +
  scale_y_continuous(limits=c(-5.5, 9)) +
  scale_x_discrete(breaks=c("diagnostic FOLR1","relapse FOLR1", "NBM FOLR1"),
                   labels=c("diagnostic", "relapse", "NBM")) +
  ggpubr::stat_compare_means(comparisons = comparisons,
                             method="wilcox") +
  scale_fill_manual(values=labels_colors) + 
  scale_color_manual(values=labels_colors) +
  # annotate(geom = "text", y=9, x=unique(subset_folr1_long$labels), 
  #          label=rep(c("FOLR2","FOLR1"), times=3)) +
  theme_classic() +
  theme(axis.text = element_text(size=12, color="black"), 
        legend.title = element_blank())


# pdf("Figures/TARGET_AML_CBFGLIS_diagnostic_relapse_NBM_FOLR1_FOLR2_violin_plot_with_stats.pdf", height = 5, width=10)
folr_violins
# dev.off()
```



# WT1 Transcript Expression 

## Tximport for transcript level TPM

```{r}
transcriptIDmap.ref <- read.delim(file.path(PROJHOME, "0000.00.02_Reference_GeneInfo/gencode.v29_RepBase.v24.01_TranscriptLevel_IDmap_1.18.20.txt"),
                                  sep="\t") 
  # mutate_at(vars(gene_id:transcript_id), ~gsub("\\.[0-9]{1,2}", "",.))

# dim(transcriptIDmap.ref) #207826     22
head(transcriptIDmap.ref)
```

```{r}
res <- dir(file.path(SCRATCH,"jlsmith3/CBFGLIS/kallisto"), pattern=".h5",recursive = T, full.names = T)
names(res) <- gsub("^.+kallisto\\/(TAR.+)_with.+h5$","\\1", res) %>% 
  gsub("-",".", .)

# head(res)
length(res) #50

tx2gene <- dplyr::select(transcriptIDmap.ref, transcript_id, gene_id)

head(tx2gene)
# dim(tx2gene)
```

```{r}
txLvl <- tximport::tximport(files=res,
                              type="kallisto",
                              tx2gene = tx2gene,
                              txIn = TRUE,
                              txOut=TRUE,
                              ignoreAfterBar = TRUE)
```

```{r}
# str(geneLvl)
names(txLvl)
# txLvl$countsFromAbundance
# saveRDS(txLvl, "Expression_Data/TARGET_AML_CBFGLIS_CordBlood_transcriptLevel_Tximport.RDS")
```

```{r}
# lapply(c("abundance","counts","length"),
#        function(df) {type <- case_when(grepl("abund",df) ~ "TPM", grepl("count",df) ~ "scaledTPM", TRUE ~ "");
#        saveRDS(geneLvl[[df]], paste0("TARGET_AML_CBFGLIS_Models_transcriptLevel_",type,"_",df,".RDS"))})
```

```{r}
TPM.reference <- readRDS(file.path(PROJHOME,"0000.00.03_ExpressionMatrices/Kallisto_GRCh38_Gencode_v29/transcript_level/TARGET_AML_RBD_Dx_Rlps_NBM_MPN_Kallisto_Quant_TranscriptLevel_Abundance_TPM.RDS"))

dim(TPM.reference)
```

samps_all 

```{r}
# TPM.all <- 
```

## FOLR1 Expression

```{r}
FOLR1.samps <- manifest %>% 
    filter(grepl("NBM|CD34_PB|diagnostic", Time_point), 
         grepl("AML|NBM|CD34_PB", Group))  %>% 
  filter(Sample %in% colnames(TPM.reference))

dim(FOLR1.samps)
```

```{r}
FOLR1.txIDs <- transcriptIDmap.ref %>%
  filter(grepl("FOLR1", gene_name))

FOLR1.txIDs
# dir.create("FOLR1")
```

```{r}
FOLR1.sqanti <- sqanti %>% 
  filter(grepl("FOLR1", associated_gene)) %>% 
  arrange(structural_category)
  # filter(!grepl("mono-exon", ))

# FOLR1.sqanti
# head(FOLR1.sqanti)
# write.csv(FOLR1.sqanti,file.path(PROJHOME, "2017.02.15_CBF-GLIS_DEG/2020.12.23_CBFGLIS_Models/SQANTI_FOLR1_PacBio_LongRead_RNAseq.csv"), row.names = FALSE)
```

```{r}
FOLR1.tx.TPM <- TPM.reference[,FOLR1.samps$Sample] %>% 
  as.data.frame() %>% 
  rownames_to_column("transcript_id") %>% 
  inner_join(., select(FOLR1.txIDs, transcript_id, gene_name), 
             by="transcript_id") %>% 
  gather(Sample, TPM, -transcript_id, -gene_name) 
  
  
FOLR1.tx.dom <- FOLR1.tx.TPM %>% 
  group_by(Sample) %>%
  mutate(Rank=rank(TPM, ties.method = "first")) %>% 
  mutate(DominantSpecies=case_when(
    Rank==max(Rank) ~ transcript_id[Rank==max(Rank)],
    Rank!=max(Rank) ~ as.character("NotDominant"))) %>% 
  mutate_at(vars(DominantSpecies), ~case_when(
    all(TPM == 0) ~ "NotExpressed", 
    any(TPM > 0) ~ .)) %>% 
  ungroup() %>% 

  #spread() causes 4 lines per sample. Need to deduplicate
  #So make NAs to zeros and collapse the rows using sum()
  spread(transcript_id, TPM)  %>% 
  group_by(Sample) %>%
  mutate_at(vars(matches("^ENST|^PB")), ~replace(.,is.na(.),0)) %>%
  mutate_at(vars(matches("^ENST|^PB")), sum) %>%
  ungroup() %>% 
  
  #remove duplicated rows
  filter(DominantSpecies != "NotDominant") %>%
  filter(!duplicated(Sample)) %>% 
  
  #limit number of digits
  mutate_at(vars(matches("^ENST|^PB")), ~round(., digits=5)) %>% 
  
  left_join(., select(FOLR1.samps,Sample,USI:AML_Subtype,Group,Tissue),
            by="Sample") %>% 
  left_join(., select(FOLR1.txIDs, transcript_id, transcript_type), 
             by=c("DominantSpecies"="transcript_id")) %>% 
  select(Sample,USI:Tissue,gene_name,transcript_type,
         DominantSpecies, matches("^ENST"), -Rank)
  

  


# dim(FOLR1.tx.TPM)
# head(FOLR1.tx.TPM)
options(scipen = 999)
# table(FOLR1.tx.TPM$Group)


# write.csv(FOLR1.tx.dom,file.path(PROJHOME, "2017.02.15_CBF-GLIS_DEG/2020.12.23_CBFGLIS_Models/TARGET_AML_GRCh38_KallistoQuant_FOLR1_transcript_Expression.csv"), row.names = FALSE)
```



```{r}
summ.stats <-    FOLR1.tx.TPM %>% 
  left_join(., select(FOLR1.samps,Sample,USI:AML_Subtype,Group,Tissue),
            by="Sample") %>% 
  group_by(AML_Subtype,transcript_id) %>%
  summarize(Mean_TPM=mean(TPM),
            Median_TPM=median(TPM),
            SD_TPM=sd(TPM),
            Max_TPM=max(TPM)) %>%
  mutate_if(is.numeric, ~round(., digits = 2)) %>% 
  ungroup()


# summ.stats
```


```{r}
FOLR1.Table <- FOLR1.tx.dom %>% 
  group_by(AML_Subtype, DominantSpecies) %>% 
  summarise(Number_of_Samples=n()) %>% 
  ungroup() %>% 
  group_by(AML_Subtype) %>% 
  mutate(Percent_of_Samples=round(Number_of_Samples/sum(Number_of_Samples)*100, digits=2)) %>% 
  ungroup() %>% 
  arrange(AML_Subtype, desc(Percent_of_Samples)) %>% 
  # mutate(AML_Subtype= gsub("AML", "AML, NOS", AML_Subtype)) %>% 
  left_join(.,  summ.stats, by=c("DominantSpecies"="transcript_id", "AML_Subtype"))


# FOLR1.Table
# View(FOLR1.Table)
# write.csv(FOLR1.Table, file.path(PROJHOME, "2017.02.15_CBF-GLIS_DEG/2020.12.23_CBFGLIS_Models/TARGET_AML_FOLR1_Dominant_Transcripts_Table.csv"), row.names = F)
```



#Session Information

```{r}
sessionInfo()
```


#Old Code

```{r fig.width=15, fig.height=5}
HH.Targets.samples <- ggplot(filter(gsva.res.TF.df,grepl("NBM|CBFA2T3-GLIS2", Comparisons)),
       aes(x=Comparisons, y=Score, fill=Comparisons)) +
  geom_jitter(alpha=1.0, aes(color=Comparisons)) +
  geom_boxplot(alpha=0.3,outlier.shape = NA) +
  # stat_compare_means(comparisons = my_comparisons[4:5]) +
  labs(y="Enrichment Score", x="", title="Hedgehog Signalling Genes") +
  
  facet_wrap(~path, scales="free") +
  scale_fill_manual(values=cc.cultures) +
  scale_color_manual(values=cc.cultures) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 35, vjust=1, hjust=1),
        strip.text = element_text(size=14))

# pdf("Figures/TARGET_AML_CBFGLIS.pts_vs_NBM_ssGSEA_boxplot.pdf", height = 5, width = 10)
HH.Targets.samples
# dev.off()

HH.Targets.cultures <- ggplot(filter(gsva.res.TF.df, !grepl("CD34_PB|NBM|CBFA2T3-GLIS2", Comparisons)),
       aes(x=Comparisons, y=Score, fill=Comparisons)) +
  ggbeeswarm::geom_quasirandom(aes(color=Comparisons)) +
  geom_boxplot(alpha=0.3,outlier.shape = NA) +
  
  # ggrepel::geom_text_repel(aes(label=Label),
  #                          point.padding = unit(1,"mm"),
  #                          min.segment.length = unit(10,"mm"),
  #                          size=3) +
  
  labs(y="Enrichment Score", x="", title="Hedgehog Signalling Genes") +
  
  facet_wrap(~path, scale="free") +
  # stat_compare_means(comparisons = my_comparisons[-c(4:5)], method.args=list(alternative="less")) +
  scale_fill_manual(values=cc.cultures) +
  scale_color_manual(values=cc.cultures) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 35, vjust=1, hjust=1),
        strip.text = element_text(size=14), 
        plot.margin = margin(l=2.5, unit="cm"))


# pdf("TARGET_AML_CBFGLIS.models_vs_Controls_ssGSEA__boxplot.pdf", height = 8, width = 20)
HH.Targets.cultures
# dev.off()
```


```{r}
EC_Week6 <-  heatmap.samps %>%
    filter(grepl("EC_Week1_GFP_CordBlood|EC_Week6_CBFGLIS_CordBlood", Comparisons)) %>% 
    mutate(USI=Sample) %>%
    set_rownames(.$Sample) %>% 
    droplevels()

levels(EC_Week6$Comparisons)

# test1 <- gsva.res.all[grep("hsa.+tgf", rownames(gsva.res.all), ignore.case = T), EC_Week6$Sample]
# test2 <- gsva.res.all[grep("hsa.+wnt", rownames(gsva.res.all), ignore.case = T), EC_Week6$Sample]
# 
# # test1
# # test2
# res1 <- wilcox.test(x=test1[grep("CBFA2T3", names(test))], y=test1[grep("GFP", names(test))], alternative = "two.sided", exact=TRUE)
# res2 <- wilcox.test(x=test2[grep("CBFA2T3", names(test))], y=test2[grep("GFP", names(test))], alternative = "greater", exact=FALSE)
# 
# # res1
# # res2

EC_Week6_DE_GSEA <- DeGSEA::gsva_DE(gsva_matrix = gsva.res.all[grep("hsa", rownames(gsva.res.all)),EC_Week6$Sample],
                              clinData=EC_Week6,
                              col="Comparisons",
                              p.value = 0.05)

dim(EC_Week6_DE_GSEA$gene_sets)
# write.csv(EC_Week6_DE_GSEA$gene_sets,"TARGET_AML_CBFGLIS.models_vs_GFP_ssGSEA_DE_Pathways.csv", row.names = FALSE)

# EC_Week6_DE_GSEA$gene_sets %>%
#   filter(grepl("Hedge|wnt|tgf",GeneSet, ignore.case = T))
```