---
title: 'miRNA-mRNA Interactions in CBFLGIS'
author: "Jenny Smith"
date: "November 16, 2018"
output: html_document
---

#Set-up 

```{r setup}
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.align='center', fig.height=5, fig.width=8)
knitr::opts_knit$set(root.dir = '~/RNA_seq_Analysis/2018.03.21_CBF-GLIS_DEGs_Comprehensive/')
options(stringsAsFactors = FALSE)
```


```{r message = FALSE, warning=FALSE}
library(stringr)
library(magrittr)
library(ggplot2)
library(dplyr)
library(tibble)
library(tidyr)
library(multiMiR, lib.loc = "~/R/x86_64-pc-linux-gnu-library/3.5/")
getwd()
```


#Define Functions 


```{r}
reformat_ints <- function(multimir.res, gene.miR.corrs, DEGs, DEMirs, type="summary"){
  
  if(type=="summary"){
     reformatted <- multimir.res@summary %>%
       filter(all.sum >= 2 | validated.sum > 0) #maybe should skip this step. Too stringent.....
     
  }else if (type=="data"){
     reformatted <- Up.Gene.Ints@data
  }
  
  reformatted <- reformatted %>%
    inner_join(., dplyr::select(gene.miR.corrs,
                      Mir.V.21,
                      Gene, 
                      SpearmanRho, 
                      adj.P.Val_Corr=Adj.P.val),
            by=c("mature_mirna_id"="Mir.V.21",
                 "target_symbol"="Gene")) %>%
    left_join(., dplyr::select(DEGs,
                      gene,
                      logFC_Gene=logFC,
                      adj.P.Val_Gene=adj.P.Val,
                      contains("Compartment")),
            by=c("target_symbol"= "gene")) %>%
    left_join(., dplyr::select(DEMirs,
                      miR,
                      logFC_miRNA=logFC,
                      adj.P.Val_miRNA=adj.P.Val,
                      gene),
            by=c("mature_mirna_id"= "miR")) %>%
    dplyr::select(mature_mirna_acc, mature_mirna_id,logFC_miRNA,adj.P.Val_miRNA,
         target_symbol,logFC_Gene,adj.P.Val_Gene,
         SpearmanRho, adj.P.Val_Corr,
         everything())
  
  return(reformatted)
  
}
```



#Read in the Clinical Data

```{r}
CDE.1031 <- read.csv("~/reference_mapping-files/AAML1031_Target_CDEs_with_HiAR_and_PrimaryCyto_withFusionCalls_10.11.2018.csv")


CDE.1031 <- CDE.1031 %>%
  filter(!is.na(USI)) %>%
  set_rownames(.$USI)


head(CDE.1031[,1:5])
```

#Read in the DE Results

```{r}
DEMirs <- read.csv("DEMirs/TARGET_AML_1031_CBFGLIS_vs_OtherAML_DEMirs_10.29.18.csv")

dim(DEMirs)
head(DEMirs)
```


```{r}
DEGs.1031 <- read.csv("DEGs/TARGET_AML_1031_CBFGLIS_vs_OtherAML_DEGs_updateAnno.csv")

dim(DEGs.1031)
```


#Read in the Correlation Data 

```{r}
UpGenePairs <- read.csv("miRNA-mRNA_Interactions/TARGET_AML_1031_CBFGLIS_vs_OtherAML_UpGenes_DnMirs_antiCorrs.csv")

head(UpGenePairs)
dim(UpGenePairs) # 452   7
```

```{r}
DnGenePairs <- read.csv("miRNA-mRNA_Interactions/TARGET_AML_1031_CBFGLIS_vs_OtherAML_DnGene_UpMirs_antiCorrs.csv")

head(DnGenePairs)
dim(DnGenePairs)
```


#Use MultiMir to Query Databases

- default is to query validated targets 
- table = "predicted" for predicted interactions

```{r}
multimir_dbInfoVersions()
```

```{r}
#ensure using the most up to date database version
multimir_switchDBVersion(db_version = "2.2.0")
```

```{r}
#mirTarbase is still out of date. I will need to manually supplement this in  my pipeline. 
multimir_dbInfo()
```

```{r warning=FALSE, message=FALSE}
Up.Gene.Ints <- get_multimir(org="hsa", 
                           mirna = UpGenePairs$Mir.V.21,
                           target = UpGenePairs$Gene,
                           table = "all", 
                           summary = TRUE, 
                           predicted.cutoff = Inf, #no filter
                           predicted.cutoff.type = "n", 
                           use.tibble = TRUE)

```

```{r}
table(Up.Gene.Ints@data$type)
dim(Up.Gene.Ints@summary) #2203 total interactions
# saveRDS(Up.Gene.Ints, "TARGET_AML_1031_CBFGLIS_vs_OtherAMLs_UpGenes_DnMirs_multimir.RDS")
```


```{r warning=FALSE, message=FALSE}
Dn.Gene.Ints <- get_multimir(org="hsa", 
                           mirna = DnGenePairs$Mir.V.21,
                           target = DnGenePairs$Gene,
                           table = "all", 
                           summary = TRUE, 
                           predicted.cutoff = Inf, 
                           predicted.cutoff.type = "n", 
                           use.tibble = TRUE)
```

```{r}
table(Dn.Gene.Ints@data$type)
dim(Dn.Gene.Ints@summary) # 12, 139 
# saveRDS(Dn.Gene.Ints, "TARGET_AML_1031_CBFGLIS_vs_OtherAMLs_DnGenes_UpMirs_multimir.RDS")
```



#Reformat the results

```{r}
Up.Gene.Ints_fmt <- reformat_ints(multimir.res = Up.Gene.Ints, 
                                  gene.miR.corrs = UpGenePairs, 
                                  DEGs=DEGs.1031,
                                  DEMirs=DEMirs) #%>%
  # mutate(Pairs=paste(mature_mirna_id, target_symbol, sep="_"))


# head(Up.Gene.Ints_fmt) 
dim(Up.Gene.Ints_fmt) #39 interactions. 2 less than the anamiR results.... 
# write.csv(Up.Gene.Ints_fmt, "TARGET_AML_1031_CBFGLIS_vs_OtherAMLs_UpGenes_DnMirs_multimir.csv", row.names = FALSE)
```

```{r}
Up.Gene.FullData <- reformat_ints(multimir.res = Up.Gene.Ints,
                                  gene.miR.corrs = UpGenePairs,
                                  DEGs=DEGs.1031, 
                                  DEMirs = DEMirs, 
                                  type = "data")

dim(Up.Gene.FullData) # 2177   22


head(Up.Gene.FullData)
```

```{r}
Dn.Gene.Ints_fmt <- reformat_ints(multimir.res = Dn.Gene.Ints, 
                                  gene.miR.corrs = DnGenePairs, 
                                  DEGs=DEGs.1031, 
                                  DEMirs=DEMirs) #%>%
  # mutate(Pairs=paste(mature_mirna_id, target_symbol, sep="_"))



# head(Dn.Gene.Ints_fmt) 
dim(Dn.Gene.Ints_fmt) #1180 interactions. 506 less than the anamiR results.... 
# write.csv(Dn.Gene.Ints_fmt, "TARGET_AML_1031_CBFGLIS_vs_OtherAMLs_DnGenes_UpMirs_multimir.csv", row.names = FALSE)
```


Ultimately, anamiR is the superiour package for finding interactions. However, it is NOT maintained anymore and will not be updated any point... SO I may use both, but even multiMiR that is actively maintained is fairly out of date....




#Examine NCAM1 interactions 


```{r}
load(url("http://multimir.ucdenver.edu/bladder.rda"))
```

```{r}
NCAM1.ints <- get_multimir(org="hsa", 
                           # mirna = NCAM1$Mir.V.21,
                           mirna = "hsa-miR-378a-5p",
                           target = "ENSG00000149294",
                           table = "all", 
                           summary = TRUE, 
                           predicted.cutoff = 500000, 
                           predicted.cutoff.type = "n", 
                           use.tibble = TRUE)


```

```{r}
table(NCAM1.ints@data$type)
```

```{r}
dim(NCAM1.ints@data)

head(NCAM1.ints@data)
write.csv(NCAM1.ints@data, "~/test.csv", row.names = FALSE)
```

```{r}
quantile(NCAM1.ints@data$score, na.rm=TRUE)
```

```{r}
NCAM1.ints@summary
```

```{r}
corr <- filter(UpGenePairs.all, Gene == "NCAM1") %>%
  dplyr::select(Adj.P.val.Corr=Adj.P.val, everything())

head(corr)
range(corr$Adj.P.val.Corr)
range(corr$SpearmanRho)
```

```{r}
NCAM1.ints.DE <- NCAM1.ints@summary %>%
  filter(mature_mirna_acc %in% DEMirs$MIMAT) %>%
  inner_join(., DEMirs, by=c("mature_mirna_acc"="MIMAT")) %>%
  left_join(., corr, by=c("mature_mirna_acc"="MIMAT")) %>%
  dplyr::select(mature_mirna_acc, mature_mirna_id, target_symbol, logFC, adj.P.Val,SpearmanRho, Adj.P.val.Corr, everything())


dim(NCAM1.ints.DE)
head(NCAM1.ints.DE)
write.csv(NCAM1.ints.DE, "TARGET_AML_CBFGLIS_NCAM1_miRNA.mRNA.csv", row.names = FALSE)
```


#Session Information 

```{r}
sessionInfo()
```

