---
title: "CBFGLIS and M7 Morphology DE Analysis"
author: "Jenny Smith"
date: "December 17, 2018"
output:
  html_document: default
  pdf_document: default
  word_document: default
---



```{r setup}
library(knitr)
knitr::opts_chunk$set(tidy.opts=list(width.cutoff=50),tidy=TRUE, fig.align='center', 
                      fig.height=10, fig.width=10)
knitr::opts_knit$set(root.dir = '~/RNA_seq_Analysis/2018.03.21_CBF-GLIS_DEGs_Comprehensive/')
options(stringsAsFactors = FALSE)
```


```{r message = FALSE, warning=FALSE}
library(stringr)
library(magrittr)
library(ggplot2)
library(gridExtra)
library(dplyr)
library(tibble)
library(tidyr)
getwd()
```

```{r}
source("~/scripts/RNAseq_Analysis/DifferentialExpn_PathwayAnalysis/DifferentialExpressionPipeline_7.05.18.r")
```


#The Requests

From Thomas Mercher

1-CBFA2T3-GLIS2 (all patients) vs. All other pediatric AML (*DONE*)
2-CBFA2T3-GLIS2 AMKL  patients vs. All other pediatric AML except the CBFA2T3-GLIS2 patients with other AML
3-CBFA2T3-GLIS2 patients with other AML vs. All other pediatric AML except the CBFA2T3-GLIS2 patients with AMKL
4-CBFA2T3-GLIS2 AMKL  patients vs. CBFA2T3-GLIS2 patients with other AML

and if you have…
5-CBFA2T3-GLIS2 AMKL  patients vs. Normal progenitors (are these adult bone marrow or cord blood normal progenitors?)
6-CBFA2T3-GLIS2 AML  patients vs. Normal progenitors



#Read in the Clinical Data

```{r}
CDE <- read.csv("~/reference_mapping-files/TARGET_AML_0531_1031_merged_CDEs_10.29.18.csv") %>%
  mutate(CBFA2T3.GLIS2=gsub("Intermediate","Yes",CBFA2T3.GLIS2))

head(CDE[,1:5])
dim(CDE)
```

```{r}
table(CDE$M7_AML, CDE$CBFA2T3.GLIS2)
```


```{r}
#to define columns per the request, any patients to be excluded will be coded as "unknown" simply so they are fitlered out of the DE analysis

CDE <- CDE %>%
  mutate(M7.CBFGLIS=case_when(
    M7_AML == "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "M7.CBFGLIS", 
    M7_AML != "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "Unknown", #so that it is filtered out. Non-M7 CBFGLIS excluded from this analysis. 
    TRUE ~ CBFA2T3.GLIS2), 
    
 nonM7.CBFGLIS= case_when(
    M7_AML == "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "Unknown", #so that it is filtered out. M7 CBFGLIS excluded from this analysis.
    M7_AML == "Unknown" & CBFA2T3.GLIS2 == "Yes" ~ "Unknown", #there are 5 CBFLGIS who have unknown morphology. remove in case one is actaully M7 (though unlikely...)
    M7_AML != "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "nonM7.CBFGLIS",  
    TRUE ~ CBFA2T3.GLIS2), 
 
 M7.CBFGLISvsNonM7.CBFGLIS=case_when(
    M7_AML == "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "M7", 
    M7_AML == "Unknown" & CBFA2T3.GLIS2 == "Yes" ~ "Unknown", #there are 5 CBFLGIS who have unknown morphology. remove in case one is actaully M7 (though unlikely...)
    M7_AML != "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "No",  
    TRUE ~ "Unknown"), 
 
 M7.CBFGLIS.Only=case_when(
    M7_AML == "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "M7.CBFGLIS", 
    M7_AML != "Yes" & CBFA2T3.GLIS2 == "Yes" ~ "Unknown", #so that it is filtered out. Non-M7 CBFGLIS excluded from this analysis. 
    TRUE ~ "Unknown")) %>%
  
  filter(!is.na(TARGET.USI.1)) %>%
  set_rownames(.$TARGET.USI.1)
  

dim(CDE)
```

```{r}
sapply(CDE[,177:180], table)
```



#Read in the Expression Data

```{r}
cts <- read.csv("~/RNA_seq_Analysis/0000.00.03_Expression_Matrices/TARGET_AML_AAML1031_dupGenesRemoved_FractionalCounts.csv", row.names = 1)
colnames(cts) <- str_split_fixed(colnames(cts), "\\.", n=2)[,1]

head(cts[,1:5])
```



#Differential Expression Analysis

## versus Other AMLs

```{r message=FALSE}
pdf("TARGET_AML_1031_DEGs_TMercher_Heatmaps.pdf", height = 10, width = 10)
DEGs <- lapply(colnames(CDE)[177:179],function(x) twoGroups_DEGs( 
                    clinData=CDE,
                    expnData = cts,
                    col=x,
                    ref="No",
                    logCPM=FALSE,
                    BM=FALSE,
                    GOI=NULL,
                    anno=FALSE,
                    method="ward.D2"))

dev.off()

names(DEGs) <- colnames(CDE)[177:179]
saveRDS(DEGs, "TARGET_AML_1031_CBFGLIS_DE_TM.RDS")
```

```{r}
summary(DEGs)
sapply(DEGs, function(x) table(x$phenovector))
```

```{r}
M7.DEGs <- DEGs$M7.CBFGLIS
nonM7.DEGs <- DEGs$nonM7.CBFGLIS
M7vNon.GLIS <- DEGs$M7.CBFGLISvsNonM7.CBFGLIS
```

```{r}
M7.DEGs.df <- extract_DEGs(M7.DEGs)
nonM7.DEGs.df <- extract_DEGs(nonM7.DEGs)
M7vNon.GLIS.df <- extract_DEGs(M7vNon.GLIS)

dim(M7.DEGs.df) #3035    8
dim(nonM7.DEGs.df) #3665    8
dim(M7vNon.GLIS.df) #293 8

# write.csv(M7.DEGs.df, "TARGET_AML_1031_M7.CBFGLIS_vs_OtherAMLs_12.18.18.csv", row.names = FALSE)
# write.csv(nonM7.DEGs.df, "TARGET_AML_1031_nonM7.CBFGLIS_vs_OtherAMLs_12.18.18.csv",row.names = FALSE)
# write.csv(M7vNon.GLIS.df, "TARGET_AML_1031_M7.CBFGLIS_vs_nonM7.CBFGLIS_12.18.18.csv",row.names = FALSE)
```


```{r fig.height=10, fig.width=14}
grid.arrange(extract_PCA(M7.DEGs),
    extract_MDS(M7.DEGs),
    
    extract_PCA(nonM7.DEGs),
    extract_MDS(nonM7.DEGs),
    
    extract_PCA(M7vNon.GLIS),
    extract_MDS(M7vNon.GLIS),ncol=2)
```


##Examine FCs of the M7.CBFGLIS to nonM7.CBFGLIS

```{r fig.height=5, fig.width=5}
# df <- M7.DEGs.df %>%
#   inner_join(., nonM7.DEGs.df, by="gene")
df <- data.frame(logFC=M7.DEGs$DE$eBayesFit$coefficients[,1]) %>%
      rownames_to_column("Gene") %>%
  inner_join(., data.frame(logFC=nonM7.DEGs$DE$eBayesFit$coefficients[,1], 
                           Gene=names(nonM7.DEGs$DE$eBayesFit$coefficients[,1])), by="Gene")

ggplot(df, aes(x=logFC.x, y=logFC.y)) +
  geom_point() +
  geom_smooth(method="lm",se = TRUE) +
  theme_bw()
```

```{r}
mod <- lm(logFC.x ~ logFC.y, data = df)
summary(mod)
```



## versus NBMs

```{r}
# pdf("TARGET_AML_1031_M7.CBFGLSI_vs_NBM_TMercher_Heatmaps.pdf", height = 10, width = 10)
DEGs.NBM <- twoGroups_DEGs( 
                    clinData=CDE,
                    expnData = cts,
                    col="M7.CBFGLIS.Only",
                    ref="No",
                    logCPM=FALSE,
                    BM=TRUE,
                    GOI=NULL,
                    anno=FALSE,
                    method="ward.D2")
# dev.off()
```

```{r}
DEGs.NBM.df <- extract_DEGs(DEGs.NBM)


dim(DEGs.NBM.df) #5213    8
# write.csv(DEGs.NBM.df,"TARGET_AML_1031_M7.CBFGLIS_vs_NBM_12.18.18.csv", row.names = FALSE)
```

```{r}
dim(DEGs.NBM$DE$Voom$E) #18,224 genes
```

```{r fig.height=5, fig.width=7}
grid.arrange(extract_PCA(DEGs.NBM),
    extract_MDS(DEGs.NBM))
```



#Compare Human DEGs to the Mouse Model DEGs 


```{r}

```





#Session Information 

```{r}
sessionInfo()
```

